---
title: "Xenium Breast 5K: cellAdmix Membrane Scoring"
output:
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---

```{r setup, include=FALSE}
extra_lib <- Sys.getenv("CELLADMIX_R_LIB", "")
if (nzchar(extra_lib)) {
  .libPaths(c(extra_lib, .libPaths()))
}

library(cellAdmixCore)
library(ggplot2)
library(cowplot)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE,
  fig.align = "center", dpi = 96, fig.retina = 2)
par(mar = c(3.5, 3.5, 2.0, 0.5), mgp = c(2, 0.65, 0), cex = 2/3)
```

## Overview

This notebook starts from the curated breast 5K annotations shipped with the
example, then runs the main cellAdmix workflow on the full
Xenium bundle: fit molecule-level factors, infer membrane-supported admixture
patterns, remove molecules matching those correction rules, and inspect the
effect of cleanup.

The workflow is:

```{r api-overview, eval=FALSE}
ds <- cellAdmix(bundle_dir, output_dir = output_dir, annotation = cell_annotation)
fit <- ds$fit(nmf_variant = "invsqrt_kl")  # membrane scoring pairs with invsqrt_kl

membrane_score <- fit$score_membrane()
membrane_rules <- membrane_score$rules()
membrane_correction <- membrane_score$correct(rules = membrane_rules)
```

## Inputs

Place the Xenium output bundle in a local folder named `data` (this example
uses a Xenium Prime 5K breast cancer run). The curated annotations ship with
this example under `annotations`: cell-type labels in `annotation.csv.gz`
and coarse tissue-domain labels in `domain_annotation.csv.gz`, produced by a
clustering/marker annotation workflow on the same bundle.

This 5K panel uses higher cell-complexity QC thresholds in the annotation and
cell-state diagnostic steps than the smaller pancreas 377-gene example:
`min_molecules = 30` and `min_genes = 15`. With the default lower thresholds,
many very small low-complexity segmented objects formed detached UMAP islands.
These thresholds are practical choices for this dataset, not universal defaults.

```{r inputs}
bundle_dir <- "data"
output_dir <- "out"
annotation_file <- file.path("annotations", "annotation.csv.gz")
domain_file <- file.path("annotations", "domain_annotation.csv.gz")

annotation <- read.csv(annotation_file, stringsAsFactors = FALSE)
domain_annotation <- read.csv(domain_file, stringsAsFactors = FALSE)

table(annotation$merged_annotation)
table(domain_annotation$domain_label)
```

The cell-type annotation is passed as a named vector. Cells labeled
`Ambiguous / low-quality` are retained in the side plots, but are excluded from
the active fit/scoring annotation so they are not treated as a biological source
or target class. The domain annotation is kept in R memory for spatial plots and
diagnostic DE comparisons; it is not used to fit factors or score admixture.

```{r construct-celladmix}
cell_annotation <- setNames(annotation$merged_annotation, annotation$cell_id)
cell_annotation <- cell_annotation[cell_annotation != "Ambiguous / low-quality"]

ds <- cellAdmix(bundle_dir, output_dir = output_dir, annotation = cell_annotation,
  num_threads = 10)

ds
```

## Fit Factors

The fit uses the `invsqrt_kl` NMF variant — the recommended pairing for
membrane scoring on stained data (see
[docs/benchmarks.md](../../docs/benchmarks.md)) — with automatic rank based
on the number of annotated cell types, multiple NMF restarts, and
gene-loading molecule scoring.

```{r fit-default}
fit <- ds$fit(nmf_variant = "invsqrt_kl", verbose = TRUE)

fit
```

Top factor loadings are the first-pass interpretation of each factor.

```{r factor-loadings, fig.width=8.5, fig.height=5.0}
fit$plot_loadings()
```

The stability plot summarizes how reproducible the factors were across NMF
starts. High-importance, high-stability factors are usually easiest to
interpret.

The stability plot shows the matched gene-ownership correlation of each factor
across independent NMF restarts (near zero for unrelated factors, above 0.3
for reproducibly re-found ones). Factor importance (the x-axis, echoed by dot size) is the share of all assigned molecules that carry the factor's label, so large dots mark factors that dominate the tissue-wide signal.

```{r factor-stability, fig.width=6.2, fig.height=4.2}
fit$plot_stability()
```

## Admixture Audit

Before correcting anything, the audit estimates each ordered cell-type
pair's admixture rate — the percent of the target type's molecules that
leaked in from the source — using only the spatial exposure structure of
the data. It frames what the cleanup below should fix, and the same audit
object verifies the correction afterwards.

```{r audit-map, fig.width=8.6, fig.height=6.4}
audit <- fit$audit_admixture()
audit$plot_map()
```

## Tissue Context

For the side visualizations below, merge cell-level factor summaries with the
provisional cell-type and domain annotations.

```{r cell-spatial-data}
cell_factors <- fit$cell_factors()
cell_spatial <- merge(cell_factors,
  annotation[, c("cell_id", "merged_annotation", "cluster_label")],
  by = "cell_id", all.x = FALSE, sort = FALSE)
cell_spatial <- merge(cell_spatial,
  domain_annotation[, c("cell_id", "domain_label")],
  by = "cell_id", all.x = FALSE, sort = FALSE)
cell_spatial$domain_label <- factor(cell_spatial$domain_label,
  levels = c("immune_rich", "tumor_epithelial"))
```

```{r spatial-celltypes, fig.width=6.8, fig.height=5.4}
celladmix_plot_spatial(cell_spatial, color_by = "merged_annotation",
  point_size = 0.06) +
  ggtitle("Cell type annotation")
```

```{r spatial-domains-factors, fig.width=12.0, fig.height=5.2}
cowplot::plot_grid(plotlist = list(
  celladmix_plot_spatial(cell_spatial, color_by = "domain_label",
    point_size = 0.06) +
    ggtitle("Coarse tissue domains"),
  celladmix_plot_spatial(cell_spatial, color_by = "dominant_factor",
    point_size = 0.06) +
    ggtitle("Dominant cellAdmix factor")
), ncol = 2, align = "hv", axis = "tblr")
```

## Diagnostic Setup

Correction is run for all annotated cell types. To keep the follow-up plots
readable, the diagnostic DE panels focus on the two large populations whose
dominant admixture sources membrane scoring calls well — epithelial cells (which
receive heavy immune leakage) and myeloid cells (which receive epithelial
leakage). The stain-weak stromal/vascular types, whose admixture the
current corrections cannot handle safely, are examined separately in the
bridge section below. The marker sets highlight likely source signatures
that may appear as admixture in the target cells.

```{r diagnostic-marker-sets}
diagnostic_cell_types <- c("Epithelial", "Myeloid")

epithelial_tumor_markers <- c("EPCAM", "KRT8", "KRT18", "KRT19", "MUC1",
  "GATA3", "CA12", "XBP1", "ERBB2")
lymphoid_myeloid_markers <- c("TRAC", "TRBC1", "CD3D", "CD3E", "CD8A",
  "IL7R", "MS4A1", "CD79A", "AIF1", "C1QC", "MS4A6A", "ITGAX")
source_marker_sets <- list(
  "epithelial/tumor source" = epithelial_tumor_markers,
  "immune source" = lymphoid_myeloid_markers
)
```

The baseline DE comparison contrasts the tumor-epithelial domain against the
immune-rich domain within each diagnostic cell type. This is an illustrative
diagnostic, not a curated biological benchmark.

```{r diagnostic-de-original}
original_counts <- fit$counts()
diagnostic_de_original <- celladmix_de_by_group(original_counts, cell_spatial,
  subset_by = "merged_annotation", subsets = diagnostic_cell_types,
  group_by = "domain_label", contrast = c("tumor_epithelial", "immune_rich"))
```

```{r baseline-volcano, fig.width=11.0, fig.height=4.8}
baseline_volcano_plots <- lapply(diagnostic_cell_types, function(target_type) {
  de <- diagnostic_de_original[[target_type]]
  celladmix_plot_volcano(de, markers = source_marker_sets,
    title = paste(target_type, "before cleanup"),
    subtitle = sprintf("logFC = tumor_epithelial / immune_rich; n = %d / %d",
      unique(de$n_group_a), unique(de$n_group_b)))
})
cowplot::plot_grid(plotlist = baseline_volcano_plots,
  ncol = length(baseline_volcano_plots), align = "hv", axis = "tblr")
```

## Membrane Scoring

Membrane scoring asks whether target-cell molecules assigned to a factor are
enriched along membrane-stain paths facing a candidate source cell type.

```{r score-settings}
p_thresh <- 0.1
score_pairs_width <- 9.6
score_pairs_height <- max(6, ceiling(fit$rank / 2) * 4.9)
```

```{r membrane-score}
membrane_score <- fit$score_membrane()
```

The factor annotation identifies likely factor source cell types and significant
target cell types. Correction rules say which factor/target combinations will
be removed.

```{r membrane-rules}
membrane_annotation <- membrane_score$annotation(p_thresh = p_thresh)
membrane_rules <- membrane_score$rules(p_thresh = p_thresh)

head(membrane_rules[, c("factor", "source_cell_type", "target_cell_type", "p_value")])
```

The heatmap shows factor-by-cell-type evidence, with symbols marking inferred
source and cleanup target calls.

```{r membrane-heatmap, fig.width=8.5, fig.height=5.4}
membrane_score$plot_heatmap(p_thresh = p_thresh)
```

The pair plots show one factor at a time. The top margin shows source evidence;
the main heatmap shows target significance for source-target cell-type pairs.

```{r membrane-score-pairs, fig.width=score_pairs_width, fig.height=score_pairs_height}
membrane_score$plot_pairs()
```

## Example Cells

These examples show the molecule-level interpretation for selected diagnostic
cells before cleanup. The examples are chosen from cells with strong
rule-supported membrane-score evidence. Molecule colors show the native factor
set in blue, the top putative admixture factor in red, and other non-native
factors in orange. Source-marker molecules are drawn slightly larger.

```{r membrane-example-overlays, fig.width=10.0, fig.height=10.0}
example_cells <- membrane_score$examples(rules = membrane_rules,
  score_annotation = membrane_annotation, cell_data = cell_spatial,
  targets = diagnostic_cell_types, n_per_target = 4)

for (cell_type in diagnostic_cell_types) {
  type_examples <- example_cells[example_cells$target_cell_type == cell_type, , drop = FALSE]
  print(membrane_score$plot_examples(type_examples,
    score_annotation = membrane_annotation, cell_data = cell_spatial,
    markers = unique(unlist(source_marker_sets)), ncol = 2))
}
```

## Cleanup

Apply all membrane-derived correction rules across all annotated cell types.
With the default multi-restart fit this is the molecule-vote ensemble: each
restart member is scored and vetted independently, and a molecule is removed
only when at least 30% of the members agree (3 of 10 here). The corrected
counts are then used for diagnostic plots.

```{r membrane-cleanup}
membrane_correction <- membrane_score$correct(rules = membrane_rules,
  name = "membrane_clean")
membrane_counts <- membrane_correction$counts()

membrane_correction$summary()
```

```{r membrane-diagnostic-de}
membrane_de <- celladmix_de_by_group(membrane_counts, cell_spatial,
  subset_by = "merged_annotation", subsets = diagnostic_cell_types,
  group_by = "domain_label", contrast = c("tumor_epithelial", "immune_rich"))
```

## Cleanup Diagnostics

This boxplot summarizes how many molecules were removed from modified cells.

```{r membrane-removal, fig.width=6.8, fig.height=5.2}
membrane_correction$plot_removed_molecules()
```

The audit verifies the cleanup against the same exposure measurements: the
report gives per-pair sensitivity and the own-marker false-removal rate,
and the panels show the estimated admixture remaining and the pooled
exposure profile before and after cleanup. The evaluation warns that the
correction removed about half of Fibroblast / CAF's own-marker molecules,
and that several pairs remain uncovered — both are early signs of the
stain-weak-type problem examined in the bridge section below.

```{r audit-verify, warning=TRUE, fig.width=9.6, fig.height=4.0}
report <- audit$evaluate(membrane_correction)
report$summary()

cowplot::plot_grid(
  audit$plot_remaining(list(corrected = membrane_correction)),
  audit$plot_exposure(correction = membrane_correction),
  ncol = 2, align = "hv", axis = "tblr")
```

Marker-expression plots show how source-marker expression changed after
cleanup in the two diagnostic cell types.

```{r membrane-marker-expression, fig.width=11.0, fig.height=4.2}
plots <- lapply(diagnostic_cell_types, function(cell_type) {
  cells <- cell_spatial$cell_id[cell_spatial$merged_annotation == cell_type]
  celladmix_plot_marker_expression(original_counts, membrane_counts,
    cells = cells, markers = source_marker_sets) +
    ggtitle(cell_type)
})
cowplot::plot_grid(plotlist = plots, ncol = length(plots), align = "hv", axis = "tblr")
```

Volcano plots show the diagnostic domain comparison before and after cleanup.

```{r membrane-volcano, fig.width=11.0, fig.height=4.8}
for (target_type in diagnostic_cell_types) {
  p_before <- celladmix_plot_volcano(diagnostic_de_original[[target_type]],
    markers = source_marker_sets, title = paste(target_type, "before cleanup")) +
    theme(legend.position = "none")
  p_after <- celladmix_plot_volcano(membrane_de[[target_type]],
    markers = source_marker_sets, title = paste(target_type, "after membrane cleanup"))
  print(cowplot::plot_grid(plotlist = list(p_before, p_after),
    ncol = 2, align = "hv", axis = "tblr"))
}
```

The p-value shift plots compare signed DE significance before and after cleanup.
Genes below the diagonal are less significant after correction.

```{r membrane-pvalue-shift, fig.width=11.0, fig.height=5.0}
plots <- lapply(diagnostic_cell_types, function(target_type) {
  celladmix_plot_de_shift(diagnostic_de_original[[target_type]], membrane_de[[target_type]],
    markers = source_marker_sets, title = paste("Membrane", target_type, "p-value shift"))
})
cowplot::plot_grid(plotlist = plots, ncol = length(plots), align = "hv", axis = "tblr")
```

## Bridge Scoring and Its Limits on Stain-Weak Targets

The audit warnings above flag large pairs that membrane scoring issued no
rule for — most prominently endothelial leakage into epithelial cells
(~888k estimated molecules) and the reciprocal epithelial leakage into
endothelial and fibroblast cells. Membrane scoring depends on stain-path contrast along
molecule-to-centroid lines, which is weak for thin, elongated cell types
such as endothelium and fibroblasts. Bridge scoring uses molecule
adjacency instead of the stain image and pairs best with the `ls_nmf`
factorization (see [docs/benchmarks.md](../../docs/benchmarks.md)). It
detects these pairs decisively — but, as shown below, it cannot correct
them safely on this dataset, and the reason is instructive.

```{r bridge-score}
fit_ls <- ds$fit()
bridge_score <- fit_ls$score_bridge()
bridge_rules <- bridge_score$rules(p_thresh = p_thresh)

stain_weak_types <- c("Endothelial", "Fibroblast / CAF")
head(bridge_rules[bridge_rules$keep &
  bridge_rules$target_cell_type %in% stain_weak_types,
  c("factor", "source_cell_type", "target_cell_type", "p_value")])
```

Applying the bridge rules, however, exposes a factorization-level failure
mode. The correction summary and the over-removal warnings show that
endothelial and fibroblast cells lose essentially *all* of their molecules:

```{r bridge-cleanup, warning=TRUE}
bridge_correction <- bridge_score$correct(rules = bridge_rules,
  name = "bridge_clean")

bridge_correction$summary()
```

The mechanism is visible in the factor composition. The rank-8 `ls_nmf`
fit collapses to a few active factors, none of them native to the small
endothelial and fibroblast populations — their molecules are carried
almost entirely by the epithelial- and immune-dominated factors:

```{r bridge-factor-composition}
ls_cells <- fit_ls$cell_factors()
ls_types <- cell_annotation[as.character(ls_cells$cell_id)]
fraction_cols <- grep("factor_\\d+_fraction", names(ls_cells), value = TRUE)
round(as.matrix(aggregate(ls_cells[, fraction_cols],
  by = list(cell_type = ls_types), FUN = mean)[, -1]), 2)
```

Each bridge rule is individually legitimate — endothelial cells genuinely
carry heavy epithelial and immune leakage, and each rule passes the
native-factor check against cells distant from *its own* source. But with
no native factor to hold the genuinely endothelial molecules, the union
of the rules covers the type's entire factor spectrum, and the correction
erases the cell type instead of cleaning it. The audit's per-type checks
flag this severe over-removal, which aggregate metrics alone would miss —
in aggregate the bridge correction looks competitive:

```{r bridge-verify, warning=TRUE, fig.width=9.6, fig.height=4.0}
bridge_report <- audit$evaluate(bridge_correction)
bridge_report$summary()

cowplot::plot_grid(
  audit$plot_remaining(list(
    `membrane\n(invsqrt_kl)` = membrane_correction,
    `bridge\n(ls_nmf)` = bridge_correction)),
  audit$plot_exposure(correction = bridge_correction),
  ncol = 2, align = "hv", axis = "tblr")
```

The bottom line for this dataset: membrane scoring cannot see the
epithelial leakage into these thin, stain-weak cell types, and bridge
scoring sees it but cannot separate it, because the factorization never
formed native endothelial or fibroblast factors. Correcting these pairs
safely requires factorization-level remedies (a rank that resolves the
small types, or anchor-based factor recovery — see
[docs/benchmarks.md](../../docs/benchmarks.md)). Downstream analysis in
this notebook therefore proceeds with the membrane correction, and the
endothelial and fibroblast admixture flagged by the audit should be kept
in mind when interpreting those cell types.

## Cell-State UMAP Before and After Cleanup

As a compact global diagnostic, compute cell-state UMAPs from sparse cell-level
counts before and after the membrane cleanup (the correction retained for
downstream use here — see the bridge section above for why the bridge
correction is not applied). The two UMAPs are independent embeddings, so
the exact coordinates are not meant to align; the comparison asks whether the
corrected counts produce clearer annotation separation.

```{r cell-state-umap-data}
state_cells_max <- 5000
original_state <- ds$cell_state_umap(cells_max = state_cells_max,
  min_molecules = 30, min_genes = 15, verbose = TRUE)
corrected_state <- membrane_correction$cell_state_umap(cells_max = state_cells_max,
  min_molecules = 30, min_genes = 15, verbose = TRUE)
cell_type_levels <- sort(unique(cell_annotation))
cell_type_palette <- setNames(grDevices::hcl.colors(length(cell_type_levels), "Dark 3"),
  cell_type_levels)
original_state$cell_type <- factor(original_state$cell_type, levels = cell_type_levels)
corrected_state$cell_type <- factor(corrected_state$cell_type, levels = cell_type_levels)
```

```{r cell-state-umap-plot, fig.width=11.0, fig.height=5.4}
state_umap_limits <- function(...) {
  frames <- list(...)
  x <- unlist(lapply(frames, `[[`, "umap_1"))
  y <- unlist(lapply(frames, `[[`, "umap_2"))
  span <- max(diff(range(x, na.rm = TRUE)), diff(range(y, na.rm = TRUE)))
  xmid <- mean(range(x, na.rm = TRUE))
  ymid <- mean(range(y, na.rm = TRUE))
  pad <- span * 0.04
  list(xlim = xmid + c(-1, 1) * (span / 2 + pad),
    ylim = ymid + c(-1, 1) * (span / 2 + pad))
}

plot_state_umap <- function(df, title, limits) {
  ggplot(df, aes(umap_1, umap_2, color = cell_type)) +
    geom_point(size = 0.55, alpha = 0.75) +
    coord_equal(xlim = limits$xlim, ylim = limits$ylim) +
    scale_color_manual(values = cell_type_palette, drop = FALSE) +
    labs(title = title, x = "UMAP 1", y = "UMAP 2", color = "Cell type") +
    theme_classic(base_size = 10) +
    guides(color = guide_legend(override.aes = list(size = 3.2, alpha = 1))) +
    theme(legend.position = "right")
}

limits <- state_umap_limits(original_state, corrected_state)
cowplot::plot_grid(plotlist = list(
  plot_state_umap(original_state, "Original counts", limits),
  plot_state_umap(corrected_state, "Membrane-corrected counts", limits)
), ncol = 2, align = "hv", axis = "tblr")
```
