---
title: "Xenium Breast 5K Medium Crop: cellAdmix Membrane Scoring"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
extra_lib <- Sys.getenv("CELLADMIX_R_LIB", "")
if (nzchar(extra_lib)) {
  .libPaths(c(extra_lib, .libPaths()))
}

library(cellAdmixCore)
library(ggplot2)
library(cowplot)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE,
  fig.align = "center", dpi = 192)
par(mar = c(3.5, 3.5, 2.0, 0.5), mgp = c(2, 0.65, 0), cex = 2/3)
```

## Overview

This is a medium-crop version of the breast 5K membrane-scoring notebook, meant
for fast iteration on content, layout, and diagnostics before rerunning the full
dataset. The crop is large enough to contain both immune-rich and
tumor-epithelial domains, but much smaller than the full 82M-molecule dataset.

## Inputs and Crop

The annotation notebook writes `annotations/annotation.csv.gz` and
`annotations/domain_annotation.csv.gz`. We keep the same annotations here but
construct a cropped cellAdmix dataset using `analysis_bbox`.

For this 5K panel we use stricter cell-complexity thresholds in the annotation
and cell-state UMAP diagnostics: `min_molecules = 30` and `min_genes = 15`.
The lower defaults produced many small low-complexity UMAP islands in this
dataset.

```{r inputs}
bundle_dir <- "data"
output_dir <- "out_medium_crop"
annotation_file <- file.path("annotations", "annotation.csv.gz")
domain_file <- file.path("annotations", "domain_annotation.csv.gz")
analysis_bbox <- c(5000, 7000, 4500, 6500)

annotation <- read.csv(annotation_file, stringsAsFactors = FALSE)
domain_annotation <- read.csv(domain_file, stringsAsFactors = FALSE)

in_crop <- annotation$cell_id %in% domain_annotation$cell_id[
  domain_annotation$x >= analysis_bbox[[1]] &
    domain_annotation$x <= analysis_bbox[[2]] &
    domain_annotation$y >= analysis_bbox[[3]] &
    domain_annotation$y <= analysis_bbox[[4]]
]
table(annotation$merged_annotation[in_crop])
table(domain_annotation$domain_label[
  domain_annotation$x >= analysis_bbox[[1]] &
    domain_annotation$x <= analysis_bbox[[2]] &
    domain_annotation$y >= analysis_bbox[[3]] &
    domain_annotation$y <= analysis_bbox[[4]]
])
```

Cells labeled `Ambiguous / low-quality` are retained in side plots but excluded
from the active fit/scoring annotation.

```{r construct-celladmix}
cell_annotation <- setNames(annotation$merged_annotation, annotation$cell_id)
cell_annotation <- cell_annotation[cell_annotation != "Ambiguous / low-quality"]

ds <- cellAdmix(bundle_dir, output_dir = output_dir, annotation = cell_annotation,
  num_threads = 10, analysis_bbox = analysis_bbox)

ds
```

## Fit Factors

The fit uses the `invsqrt_kl` NMF variant, the recommended pairing for
membrane scoring on stained data (see
[docs/benchmarks.md](../../docs/benchmarks.md)).

```{r fit-default}
fit <- ds$fit(nmf_variant = "invsqrt_kl", verbose = TRUE)

fit
```

```{r factor-loadings, fig.width=8.5, fig.height=5.0}
fit$plot_loadings()
```

The stability plot shows the matched gene-ownership correlation of each factor
across independent NMF restarts (near zero for unrelated factors, above 0.3
for reproducibly re-found ones). Factor importance (the x-axis, echoed by dot size) is the share of all assigned molecules that carry the factor's label, so large dots mark factors that dominate the tissue-wide signal.

```{r factor-stability, fig.width=6.2, fig.height=4.2}
fit$plot_stability()
```

## Admixture Audit

Before correcting anything, the audit estimates how many molecules leaked
between each ordered cell-type pair within the crop, using only the
spatial exposure structure of the data (cell text: estimated leaked
molecules, a conservative lower bound).

```{r audit-map, fig.width=8.6, fig.height=6.4}
audit <- fit$audit_admixture()
audit$plot_map()
```

## Tissue Context

```{r cell-spatial-data}
cell_factors <- fit$cell_factors()
cell_spatial <- merge(cell_factors,
  annotation[, c("cell_id", "merged_annotation", "cluster_label")],
  by = "cell_id", all.x = FALSE, sort = FALSE)
cell_spatial <- merge(cell_spatial,
  domain_annotation[, c("cell_id", "domain_label")],
  by = "cell_id", all.x = FALSE, sort = FALSE)
cell_spatial$domain_label <- factor(cell_spatial$domain_label,
  levels = c("immune_rich", "tumor_epithelial"))
```

```{r spatial-context, fig.width=14.2, fig.height=5.2}
cowplot::plot_grid(plotlist = list(
  celladmix_plot_spatial(cell_spatial, color_by = "merged_annotation",
    point_size = 0.18) +
    ggtitle("Cell type annotation"),
  celladmix_plot_spatial(cell_spatial, color_by = "domain_label",
    point_size = 0.18) +
    ggtitle("Tissue domain"),
  celladmix_plot_spatial(cell_spatial, color_by = "dominant_factor",
    point_size = 0.18) +
    ggtitle("Dominant cellAdmix factor")
), ncol = 3, align = "hv", axis = "tblr")
```

## Diagnostic Setup

```{r diagnostic-marker-sets}
diagnostic_cell_types <- c("Fibroblast / CAF", "Endothelial")

epithelial_tumor_markers <- c("EPCAM", "KRT8", "KRT18", "KRT19", "MUC1",
  "GATA3", "CA12", "XBP1", "ERBB2")
lymphoid_myeloid_markers <- c("TRAC", "TRBC1", "CD3D", "CD3E", "CD8A",
  "IL7R", "MS4A1", "CD79A", "AIF1", "C1QC", "MS4A6A", "ITGAX")
source_marker_sets <- list(
  "epithelial/tumor source" = epithelial_tumor_markers,
  "immune source" = lymphoid_myeloid_markers
)
```

```{r diagnostic-de-original}
original_counts <- fit$counts()
diagnostic_de_original <- celladmix_de_by_group(original_counts, cell_spatial,
  subset_by = "merged_annotation", subsets = diagnostic_cell_types,
  group_by = "domain_label", contrast = c("tumor_epithelial", "immune_rich"))
```

```{r baseline-volcano, fig.width=11.0, fig.height=4.8}
baseline_volcano_plots <- lapply(diagnostic_cell_types, function(target_type) {
  de <- diagnostic_de_original[[target_type]]
  celladmix_plot_volcano(de, markers = source_marker_sets,
    title = paste(target_type, "before cleanup"),
    subtitle = sprintf("logFC = tumor_epithelial / immune_rich; n = %d / %d",
      unique(de$n_group_a), unique(de$n_group_b)))
})
cowplot::plot_grid(plotlist = baseline_volcano_plots,
  ncol = length(baseline_volcano_plots), align = "hv", axis = "tblr")
```

## Membrane Scoring

This crop is small enough that a permissive rule threshold can produce broad
cleanup calls. We use a stricter `p_thresh = 0.01` here to keep the exploratory
crop diagnostics interpretable before moving back to full-dataset settings.

```{r score-settings}
p_thresh <- 0.01
score_pairs_width <- 9.6
score_pairs_height <- max(6, ceiling(fit$rank / 2) * 4.9)
```

```{r membrane-score}
membrane_score <- fit$score_membrane()
membrane_annotation <- membrane_score$annotation(p_thresh = p_thresh)
membrane_rules <- membrane_score$rules(p_thresh = p_thresh)

head(membrane_rules[, c("factor", "source_cell_type", "target_cell_type", "p_value")])
```

```{r membrane-heatmap, fig.width=8.5, fig.height=5.4}
membrane_score$plot_heatmap(p_thresh = p_thresh)
```

```{r membrane-score-pairs, fig.width=score_pairs_width, fig.height=score_pairs_height}
membrane_score$plot_pairs()
```

## Example Cells

These examples show the molecule-level interpretation for selected diagnostic
cells before cleanup. The examples are chosen from cells with strong
rule-supported membrane-score evidence. Molecule colors show the native factor
set in blue, the top putative admixture factor in red, and other non-native
factors in orange. Source-marker molecules are drawn slightly larger.

```{r membrane-example-overlays, fig.width=10.0, fig.height=10.0}
example_cells <- membrane_score$examples(rules = membrane_rules,
  score_annotation = membrane_annotation, cell_data = cell_spatial,
  targets = diagnostic_cell_types, n_per_target = 4)

for (cell_type in diagnostic_cell_types) {
  type_examples <- example_cells[example_cells$target_cell_type == cell_type, , drop = FALSE]
  print(membrane_score$plot_examples(type_examples,
    score_annotation = membrane_annotation, cell_data = cell_spatial,
    markers = unique(unlist(source_marker_sets)), ncol = 2))
}
```

## Cleanup Diagnostics

```{r membrane-cleanup}
membrane_correction <- membrane_score$correct(rules = membrane_rules,
  name = "membrane_clean_p001")
membrane_counts <- membrane_correction$counts()

membrane_correction$summary()
```

```{r membrane-diagnostic-de}
membrane_de <- celladmix_de_by_group(membrane_counts, cell_spatial,
  subset_by = "merged_annotation", subsets = diagnostic_cell_types,
  group_by = "domain_label", contrast = c("tumor_epithelial", "immune_rich"))
```

```{r membrane-removal, fig.width=6.8, fig.height=5.2}
membrane_correction$plot_removed_molecules()
```

The audit verifies the cleanup against the same exposure measurements it
used to detect the admixture.

```{r audit-verify, warning=TRUE, fig.width=9.6, fig.height=4.0}
report <- audit$evaluate(membrane_correction)
report$summary()

cowplot::plot_grid(
  audit$plot_remaining(list(corrected = membrane_correction)),
  audit$plot_exposure(correction = membrane_correction),
  ncol = 2, align = "hv", axis = "tblr")
```

```{r membrane-marker-expression, fig.width=11.0, fig.height=4.2}
plots <- lapply(diagnostic_cell_types, function(cell_type) {
  cells <- cell_spatial$cell_id[cell_spatial$merged_annotation == cell_type]
  celladmix_plot_marker_expression(original_counts, membrane_counts,
    cells = cells, markers = source_marker_sets) +
    ggtitle(cell_type)
})
cowplot::plot_grid(plotlist = plots, ncol = length(plots), align = "hv", axis = "tblr")
```

```{r membrane-volcano, fig.width=11.0, fig.height=4.8}
for (target_type in diagnostic_cell_types) {
  p_before <- celladmix_plot_volcano(diagnostic_de_original[[target_type]],
    markers = source_marker_sets, title = paste(target_type, "before cleanup")) +
    theme(legend.position = "none")
  p_after <- celladmix_plot_volcano(membrane_de[[target_type]],
    markers = source_marker_sets, title = paste(target_type, "after membrane cleanup"))
  print(cowplot::plot_grid(plotlist = list(p_before, p_after),
    ncol = 2, align = "hv", axis = "tblr"))
}
```

```{r membrane-pvalue-shift, fig.width=11.0, fig.height=5.0}
plots <- lapply(diagnostic_cell_types, function(target_type) {
  celladmix_plot_de_shift(diagnostic_de_original[[target_type]], membrane_de[[target_type]],
    markers = source_marker_sets, title = paste("Membrane", target_type, "p-value shift"))
})
cowplot::plot_grid(plotlist = plots, ncol = length(plots), align = "hv", axis = "tblr")
```

## Cell-State UMAP Before and After Cleanup

```{r cell-state-umap-data}
state_cells_max <- NULL
original_state_all <- ds$cell_state_umap(cells_max = state_cells_max,
  min_molecules = 30, min_genes = 15, verbose = TRUE)
corrected_state_all <- membrane_correction$cell_state_umap(cells_max = state_cells_max,
  min_molecules = 30, min_genes = 15, verbose = TRUE)
cell_type_levels <- sort(unique(cell_annotation))
cell_type_palette <- setNames(grDevices::hcl.colors(length(cell_type_levels), "Dark 3"),
  cell_type_levels)
original_state_all$cell_type <- factor(original_state_all$cell_type, levels = cell_type_levels)
corrected_state_all$cell_type <- factor(corrected_state_all$cell_type, levels = cell_type_levels)

shared_state_cells <- intersect(original_state_all$cell_id, corrected_state_all$cell_id)
original_state <- original_state_all[original_state_all$cell_id %in% shared_state_cells, , drop = FALSE]
corrected_state <- corrected_state_all[corrected_state_all$cell_id %in% shared_state_cells, , drop = FALSE]
original_state$cell_type <- factor(original_state$cell_type, levels = cell_type_levels)
corrected_state$cell_type <- factor(corrected_state$cell_type, levels = cell_type_levels)

state_cell_retention <- rbind(
  data.frame(panel = "original eligible", as.data.frame(table(original_state_all$cell_type))),
  data.frame(panel = "corrected eligible", as.data.frame(table(corrected_state_all$cell_type))),
  data.frame(panel = "plotted common cells", as.data.frame(table(original_state$cell_type)))
)
names(state_cell_retention) <- c("panel", "cell_type", "cells")
state_cell_retention
```

```{r cell-state-umap-plot, fig.width=11.0, fig.height=5.4}
state_umap_limits <- function(...) {
  frames <- list(...)
  x <- unlist(lapply(frames, `[[`, "umap_1"))
  y <- unlist(lapply(frames, `[[`, "umap_2"))
  span <- max(diff(range(x, na.rm = TRUE)), diff(range(y, na.rm = TRUE)))
  xmid <- mean(range(x, na.rm = TRUE))
  ymid <- mean(range(y, na.rm = TRUE))
  pad <- span * 0.04
  list(xlim = xmid + c(-1, 1) * (span / 2 + pad),
    ylim = ymid + c(-1, 1) * (span / 2 + pad))
}

plot_state_umap <- function(df, title, limits) {
  ggplot(df, aes(umap_1, umap_2, color = cell_type)) +
    geom_point(size = 0.55, alpha = 0.75) +
    coord_equal(xlim = limits$xlim, ylim = limits$ylim) +
    scale_color_manual(values = cell_type_palette, drop = FALSE) +
    labs(title = title, x = "UMAP 1", y = "UMAP 2", color = "Cell type") +
    theme_classic(base_size = 10) +
    guides(color = guide_legend(override.aes = list(size = 3.2, alpha = 1))) +
    theme(legend.position = "right")
}

limits <- state_umap_limits(original_state, corrected_state)
cowplot::plot_grid(plotlist = list(
  plot_state_umap(original_state, "Original counts", limits),
  plot_state_umap(corrected_state, "Corrected counts", limits)
), ncol = 2, align = "hv", axis = "tblr")
```
