---
title: "Xenium Breast 5K: cellAdmix Membrane Scoring"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
extra_lib <- Sys.getenv("CELLADMIX_R_LIB", "")
if (nzchar(extra_lib)) {
  .libPaths(c(extra_lib, .libPaths()))
}

library(cellAdmixCore)
library(ggplot2)
library(cowplot)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE,
  fig.align = "center", dpi = 192)
par(mar = c(3.5, 3.5, 2.0, 0.5), mgp = c(2, 0.65, 0), cex = 2/3)
```

## Overview

This notebook starts from the provisional breast 5K annotations generated in
`breast_5k_annotation.Rmd`, then runs the main cellAdmix workflow on the full
Xenium bundle: fit molecule-level factors, infer membrane-supported admixture
patterns, remove molecules matching those correction rules, and inspect the
effect of cleanup.

The workflow is:

```{r api-overview, eval=FALSE}
ds <- cellAdmix(bundle_dir, output_dir = output_dir, annotation = cell_annotation)
fit <- ds$fit()

membrane_score <- fit$score_membrane()
membrane_rules <- membrane_score$rules()
membrane_correction <- membrane_score$correct(rules = membrane_rules)
```

## Inputs

Place the Xenium output bundle in a local folder named `data`. The annotation
notebook writes two compressed files under `annotations`: cell-type labels in
`annotation.csv.gz` and coarse tissue-domain labels in
`domain_annotation.csv.gz`. The output directory `out` is shared with the
annotation notebook so the count-only store can be reused and upgraded to a
full molecule store only when the fit requires it.

This 5K panel uses higher cell-complexity QC thresholds in the annotation and
cell-state diagnostic steps than the smaller pancreas 377-gene example:
`min_molecules = 30` and `min_genes = 15`. With the default lower thresholds,
many very small low-complexity segmented objects formed detached UMAP islands.
These thresholds are practical choices for this dataset, not universal defaults.

```{r inputs}
bundle_dir <- "data"
output_dir <- "out"
nmf_variant <- Sys.getenv("CELLADMIX_NMF_VARIANT", "ls_nmf")
annotation_file <- file.path("annotations", "annotation.csv.gz")
domain_file <- file.path("annotations", "domain_annotation.csv.gz")

annotation <- read.csv(annotation_file, stringsAsFactors = FALSE)
domain_annotation <- read.csv(domain_file, stringsAsFactors = FALSE)

table(annotation$merged_annotation)
table(domain_annotation$domain_label)
```

The cell-type annotation is passed as a named vector. Cells labeled
`Ambiguous / low-quality` are retained in the side plots, but are excluded from
the active fit/scoring annotation so they are not treated as a biological source
or target class. The domain annotation is kept in R memory for spatial plots and
diagnostic DE comparisons; it is not used to fit factors or score admixture.

```{r construct-celladmix}
cell_annotation <- setNames(annotation$merged_annotation, annotation$cell_id)
cell_annotation <- cell_annotation[cell_annotation != "Ambiguous / low-quality"]

ds <- cellAdmix(bundle_dir, output_dir = output_dir, annotation = cell_annotation,
  num_threads = 10)

ds
```

## Fit Factors

The default fit uses inverse-square-root weighted KL NMF, automatic rank based
on the number of annotated cell types, multiple NMF starts, and gene-loading
molecule scoring. To render a direct KL comparison, set
`CELLADMIX_NMF_VARIANT=kl` before executing the notebook.

```{r fit-default}
fit <- ds$fit(nmf_variant = nmf_variant, verbose = TRUE)

fit
```

Top factor loadings are the first-pass interpretation of each factor.

```{r factor-loadings, fig.width=8.5, fig.height=5.0}
fit$plot_loadings()
```

The stability plot summarizes how reproducible the factors were across NMF
starts. High-importance, high-stability factors are usually easiest to
interpret.

The stability plot shows the matched gene-ownership correlation of each factor
across independent NMF restarts (near zero for unrelated factors, above 0.3
for reproducibly re-found ones). Factor importance (the x-axis, echoed by dot size) is the share of all assigned molecules that carry the factor's label, so large dots mark factors that dominate the tissue-wide signal.

```{r factor-stability, fig.width=6.2, fig.height=4.2}
fit$plot_stability()
```

## Tissue Context

For the side visualizations below, merge cell-level factor summaries with the
provisional cell-type and domain annotations.

```{r cell-spatial-data}
cell_factors <- fit$cell_factors()
cell_spatial <- merge(cell_factors,
  annotation[, c("cell_id", "merged_annotation", "cluster_label")],
  by = "cell_id", all.x = FALSE, sort = FALSE)
cell_spatial <- merge(cell_spatial,
  domain_annotation[, c("cell_id", "domain_label")],
  by = "cell_id", all.x = FALSE, sort = FALSE)
cell_spatial$domain_label <- factor(cell_spatial$domain_label,
  levels = c("immune_rich", "tumor_epithelial"))
```

```{r spatial-celltypes, fig.width=6.8, fig.height=5.4}
celladmix_plot_spatial(cell_spatial, color_by = "merged_annotation",
  point_size = 0.06) +
  ggtitle("Cell type annotation")
```

```{r spatial-domains-factors, fig.width=12.0, fig.height=5.2}
cowplot::plot_grid(plotlist = list(
  celladmix_plot_spatial(cell_spatial, color_by = "domain_label",
    point_size = 0.06) +
    ggtitle("Coarse tissue domains"),
  celladmix_plot_spatial(cell_spatial, color_by = "dominant_factor",
    point_size = 0.06) +
    ggtitle("Dominant cellAdmix factor")
), ncol = 2, align = "hv", axis = "tblr")
```

## Diagnostic Setup

Correction is run for all annotated cell types. To keep the follow-up plots
readable, the diagnostic DE panels focus on stromal/vascular cells that are
present in both coarse domains. The marker sets highlight likely source
signatures that may appear as admixture in those target cells.

```{r diagnostic-marker-sets}
diagnostic_cell_types <- c("Fibroblast / CAF", "Endothelial")

epithelial_tumor_markers <- c("EPCAM", "KRT8", "KRT18", "KRT19", "MUC1",
  "GATA3", "CA12", "XBP1", "ERBB2")
lymphoid_myeloid_markers <- c("TRAC", "TRBC1", "CD3D", "CD3E", "CD8A",
  "IL7R", "MS4A1", "CD79A", "AIF1", "C1QC", "MS4A6A", "ITGAX")
source_marker_sets <- list(
  "epithelial/tumor source" = epithelial_tumor_markers,
  "immune source" = lymphoid_myeloid_markers
)
```

The baseline DE comparison contrasts the tumor-epithelial domain against the
immune-rich domain within each diagnostic cell type. This is an illustrative
diagnostic, not a curated biological benchmark.

```{r diagnostic-de-original}
original_counts <- fit$counts()
diagnostic_de_original <- celladmix_de_by_group(original_counts, cell_spatial,
  subset_by = "merged_annotation", subsets = diagnostic_cell_types,
  group_by = "domain_label", contrast = c("tumor_epithelial", "immune_rich"))
```

```{r baseline-volcano, fig.width=11.0, fig.height=4.8}
baseline_volcano_plots <- lapply(diagnostic_cell_types, function(target_type) {
  de <- diagnostic_de_original[[target_type]]
  celladmix_plot_volcano(de, markers = source_marker_sets,
    title = paste(target_type, "before cleanup"),
    subtitle = sprintf("logFC = tumor_epithelial / immune_rich; n = %d / %d",
      unique(de$n_group_a), unique(de$n_group_b)))
})
cowplot::plot_grid(plotlist = baseline_volcano_plots,
  ncol = length(baseline_volcano_plots), align = "hv", axis = "tblr")
```

## Membrane Scoring

Membrane scoring asks whether target-cell molecules assigned to a factor are
enriched along membrane-stain paths facing a candidate source cell type.

```{r score-settings}
p_thresh <- 0.1
score_pairs_width <- 9.6
score_pairs_height <- max(6, ceiling(fit$rank / 2) * 4.9)
```

```{r membrane-score}
membrane_score <- fit$score_membrane()
```

The factor annotation identifies likely factor source cell types and significant
target cell types. Correction rules say which factor/target combinations will
be removed.

```{r membrane-rules}
membrane_annotation <- membrane_score$annotation(p_thresh = p_thresh)
membrane_rules <- membrane_score$rules(p_thresh = p_thresh)

head(membrane_rules[, c("factor", "source_cell_type", "target_cell_type", "p_value")])
```

The heatmap shows factor-by-cell-type evidence, with symbols marking inferred
source and cleanup target calls.

```{r membrane-heatmap, fig.width=8.5, fig.height=5.4}
membrane_score$plot_heatmap(p_thresh = p_thresh)
```

The pair plots show one factor at a time. The top margin shows source evidence;
the main heatmap shows target significance for source-target cell-type pairs.

```{r membrane-score-pairs, fig.width=score_pairs_width, fig.height=score_pairs_height}
membrane_score$plot_pairs()
```

## Example Cells

These examples show the molecule-level interpretation for selected diagnostic
cells before cleanup. The examples are chosen from cells with strong
rule-supported membrane-score evidence. Molecule colors show the native factor
set in blue, the top putative admixture factor in red, and other non-native
factors in orange. Source-marker molecules are drawn slightly larger.

```{r membrane-example-overlays, fig.width=10.0, fig.height=10.0}
example_cells <- membrane_score$examples(rules = membrane_rules,
  score_annotation = membrane_annotation, cell_data = cell_spatial,
  targets = diagnostic_cell_types, n_per_target = 4)

for (cell_type in diagnostic_cell_types) {
  type_examples <- example_cells[example_cells$target_cell_type == cell_type, , drop = FALSE]
  print(membrane_score$plot_examples(type_examples,
    score_annotation = membrane_annotation, cell_data = cell_spatial,
    markers = unique(unlist(source_marker_sets)), ncol = 2))
}
```

## Cleanup

Apply all membrane-derived correction rules across all annotated cell types.
The corrected counts are then used for diagnostic plots.

```{r membrane-cleanup}
membrane_correction <- membrane_score$correct(rules = membrane_rules,
  name = "membrane_clean")
membrane_counts <- membrane_correction$counts()

membrane_correction$summary()
```

```{r membrane-diagnostic-de}
membrane_de <- celladmix_de_by_group(membrane_counts, cell_spatial,
  subset_by = "merged_annotation", subsets = diagnostic_cell_types,
  group_by = "domain_label", contrast = c("tumor_epithelial", "immune_rich"))
```

## Cleanup Diagnostics

This boxplot summarizes how many molecules were removed from modified cells.

```{r membrane-removal, fig.width=6.8, fig.height=5.2}
membrane_correction$plot_removed_molecules()
```

Marker-expression plots show how source-marker expression changed after
cleanup in the two diagnostic cell types.

```{r membrane-marker-expression, fig.width=11.0, fig.height=4.2}
plots <- lapply(diagnostic_cell_types, function(cell_type) {
  cells <- cell_spatial$cell_id[cell_spatial$merged_annotation == cell_type]
  celladmix_plot_marker_expression(original_counts, membrane_counts,
    cells = cells, markers = source_marker_sets) +
    ggtitle(cell_type)
})
cowplot::plot_grid(plotlist = plots, ncol = length(plots), align = "hv", axis = "tblr")
```

Volcano plots show the diagnostic domain comparison before and after cleanup.

```{r membrane-volcano, fig.width=11.0, fig.height=4.8}
for (target_type in diagnostic_cell_types) {
  p_before <- celladmix_plot_volcano(diagnostic_de_original[[target_type]],
    markers = source_marker_sets, title = paste(target_type, "before cleanup")) +
    theme(legend.position = "none")
  p_after <- celladmix_plot_volcano(membrane_de[[target_type]],
    markers = source_marker_sets, title = paste(target_type, "after membrane cleanup"))
  print(cowplot::plot_grid(plotlist = list(p_before, p_after),
    ncol = 2, align = "hv", axis = "tblr"))
}
```

The p-value shift plots compare signed DE significance before and after cleanup.
Genes below the diagonal are less significant after correction.

```{r membrane-pvalue-shift, fig.width=11.0, fig.height=5.0}
plots <- lapply(diagnostic_cell_types, function(target_type) {
  celladmix_plot_de_shift(diagnostic_de_original[[target_type]], membrane_de[[target_type]],
    markers = source_marker_sets, title = paste("Membrane", target_type, "p-value shift"))
})
cowplot::plot_grid(plotlist = plots, ncol = length(plots), align = "hv", axis = "tblr")
```

## Cell-State UMAP Before and After Cleanup

As a compact global diagnostic, compute cell-state UMAPs from sparse cell-level
counts before and after cleanup. The two UMAPs are independent embeddings, so
the exact coordinates are not meant to align; the comparison asks whether the
corrected counts produce clearer annotation separation.

```{r cell-state-umap-data}
state_cells_max <- 5000
original_state <- ds$cell_state_umap(cells_max = state_cells_max,
  min_molecules = 30, min_genes = 15, verbose = TRUE)
corrected_state <- membrane_correction$cell_state_umap(cells_max = state_cells_max,
  min_molecules = 30, min_genes = 15, verbose = TRUE)
cell_type_levels <- sort(unique(cell_annotation))
cell_type_palette <- setNames(grDevices::hcl.colors(length(cell_type_levels), "Dark 3"),
  cell_type_levels)
original_state$cell_type <- factor(original_state$cell_type, levels = cell_type_levels)
corrected_state$cell_type <- factor(corrected_state$cell_type, levels = cell_type_levels)
```

```{r cell-state-umap-plot, fig.width=11.0, fig.height=5.4}
state_umap_limits <- function(...) {
  frames <- list(...)
  x <- unlist(lapply(frames, `[[`, "umap_1"))
  y <- unlist(lapply(frames, `[[`, "umap_2"))
  span <- max(diff(range(x, na.rm = TRUE)), diff(range(y, na.rm = TRUE)))
  xmid <- mean(range(x, na.rm = TRUE))
  ymid <- mean(range(y, na.rm = TRUE))
  pad <- span * 0.04
  list(xlim = xmid + c(-1, 1) * (span / 2 + pad),
    ylim = ymid + c(-1, 1) * (span / 2 + pad))
}

plot_state_umap <- function(df, title, limits) {
  ggplot(df, aes(umap_1, umap_2, color = cell_type)) +
    geom_point(size = 0.55, alpha = 0.75) +
    coord_equal(xlim = limits$xlim, ylim = limits$ylim) +
    scale_color_manual(values = cell_type_palette, drop = FALSE) +
    labs(title = title, x = "UMAP 1", y = "UMAP 2", color = "Cell type") +
    theme_classic(base_size = 10) +
    guides(color = guide_legend(override.aes = list(size = 3.2, alpha = 1))) +
    theme(legend.position = "right")
}

limits <- state_umap_limits(original_state, corrected_state)
cowplot::plot_grid(plotlist = list(
  plot_state_umap(original_state, "Original counts", limits),
  plot_state_umap(corrected_state, "Corrected counts", limits)
), ncol = 2, align = "hv", axis = "tblr")
```
