---
title: "Minimal cellAdmix CosMx Example"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
library(cellAdmixCore)

example_dir <- getwd()

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", dpi = 192)
par(mar = c(3.5, 3.5, 2.0, 0.5), mgp = c(2, 0.65, 0), cex = 2/3)
```

## Overview

This notebook illustrates the basic cellAdmix workflow on a CosMx NSCLC
example. The goal is to identify molecular factors that describe local
transcript neighborhoods, annotate those factors with likely source cell types,
detect source-target admixture patterns, and produce corrected cell-by-gene
counts after removing molecules assigned to likely admixture.

At a high level, the user-facing workflow is:

```{r api-overview, eval=FALSE}
ds <- cellAdmix(molecules_file, output_dir = output_dir, annotation = cell_annotation)
fit <- ds$fit()
score <- fit$score_bridge()
rules <- score$rules()
correction <- score$correct(rules = rules)
corrected_counts <- correction$counts()
```

The constructor records where the input files and output cache live, validates
the configuration, and stores lightweight cell-level metadata. It does not load
the full molecule table into R memory. The computational work starts with
`fit()`, which builds neighborhood composition vectors, fits NMF factors, and
assigns molecules to factors. Scoring then interprets these factors as possible
source-target admixture patterns, and correction removes rule-supported
admixture molecules to produce corrected counts.

## Data Preparation

cellAdmix itself works with generic molecule tables and optional cell metadata.
The raw inputs for this example are more dataset-specific: a Giotto object with
cell metadata, a CosMx transcript CSV, and a small curated annotation table. We
keep that conversion outside the main notebook so that the vignette focuses on
the standard cellAdmix API rather than on one dataset's loading details.

To regenerate the prepared files, download the raw example files from
`http://pklab.org/peterk/cellAdmix/examples/cosmx_nsclc/` and place them in a
local folder named `data/` next to this notebook:

- `data/SMI_Giotto_Object.RData`
- `data/Lung5_Rep1_tx_file.csv.gz`
- `data/annotation_adj.csv.gz`

Then run the conversion script from the `examples/cosmx_nsclc_giotto`
directory:

`Rscript prepare_cosmx_tabular.R --input data --output prepared`

The script writes gzip-compressed tabular files under `prepared/`:

- `prepared/molecules_all.csv.gz`: one molecule per row, with spatial
  coordinates, gene identity, and assigned cell ID.
- `prepared/cell_metadata_all.csv.gz`: one row per cell, with cell type labels,
  spatial centroids, and a few diagnostic columns used later for illustration.

The conversion does not require the Giotto package to be installed. It loads the
saved Giotto object and reads only the object attributes containing cell-level
metadata and raw spatial cell centroids. The curated annotation table supplies
the cell-type labels and diagnostic region labels. The transcript CSV supplies
the molecule coordinates, genes, FOV IDs, and raw cell IDs; the script filters
out unassigned cells and negative probes, converts CosMx pixel coordinates to
microns, and writes the standard tabular molecule file consumed by cellAdmix.

The rest of the notebook starts from those prepared files. This is the same
input pattern that can be used for other tabular molecule datasets.

## Prepared Inputs

We first point the notebook at the prepared molecule and cell-metadata files,
then load the small cell metadata table into R for annotation summaries and
plotting.

```{r prepared-inputs}
molecules_file <- file.path("prepared", "molecules_all.csv.gz")
cell_metadata_file <- file.path("prepared", "cell_metadata_all.csv.gz")

cell_meta <- read.csv(cell_metadata_file, stringsAsFactors = FALSE)
rownames(cell_meta) <- cell_meta$cell

table(cell_meta$cell_type_coarse)
```

The spatial plot below is not part of the core algorithm. It is a quick check
that the prepared annotations and cell coordinates are aligned with the tissue
layout.

```{r tissue-context, fig.width=5.5, fig.height=4.5}
celladmix_plot_spatial(cell_meta, color_by = "cell_type_coarse")
```

## cellAdmix Fit

The `cellAdmix()` constructor records the input files and creates an output
directory for reusable caches and analysis results. The coarse cell annotation
is passed as a named vector, where names are cell IDs and values are cell-type
labels. The molecule CSV is not loaded into R memory at construction time; it
is streamed into a cellAdmix store as needed by downstream steps.

```{r construct-celladmix}
output_dir <- "out"
cell_annotation <- setNames(cell_meta$cell_type_coarse, cell_meta$cell)

ds <- cellAdmix(molecules_file, output_dir = output_dir, annotation = cell_annotation)

ds
```

The fit call constructs neighborhood composition vectors around molecules,
factorizes them, and assigns each molecule to a factor. By default, the rank is
chosen from the number of annotation labels, the neighborhood size is resolved
from the panel and cell sizes, NMF uses the weighted least-squares (`ls_nmf`)
formulation, molecule scoring uses gene loadings, and the number of NMF starts
matches the dataset thread setting.

```{r fit-default}
fit <- ds$fit()
score_pairs_width <- 9.6
score_pairs_height <- max(6, ceiling(fit$rank / 2) * 4.9)

fit
```

Top gene loadings help interpret each factor. These are the genes most strongly
associated with each fitted factor, so they are the first place to look for
native cell-type signatures or possible admixture signatures.

```{r factor-loadings, fig.width=8, fig.height=4.8}
fit$plot_loadings()
```

The stability plot summarizes how reproducible factors were across the random
NMF starts. Stability is the matched correlation of each factor's
gene-ownership profile against independent restarts: unrelated factors score
near zero, and values above 0.3 indicate a reproducibly re-found factor.
Factor importance (the x-axis, echoed by dot size) is the share of all assigned molecules that carry the factor's label, so large dots mark factors that dominate the tissue-wide signal.
Stable, high-importance factors are easier to interpret; factors below the
threshold are seed artifacts and should not drive correction decisions.

```{r factor-stability, fig.width=6.2, fig.height=4.2}
fit$plot_stability()
```

As another quick sanity check, we can show the dominant factor assignment for
each cell in tissue coordinates.

```{r spatial-factors, fig.width=5.5, fig.height=4.5}
cell_factors <- fit$cell_factors()
celladmix_plot_spatial(cell_factors, color_by = "dominant_factor") +
  ggplot2::ggtitle("Dominant factor by cell")
```

## Bridge Scoring

Bridge scoring asks whether molecules assigned to a factor in a target cell
type are spatially connected to plausible source cell types. This converts the
unlabeled factors into candidate source-target correction rules.

```{r bridge-score}
bridge <- fit$score_bridge()
```

The score object is then converted into correction rules. Each row indicates a
factor, its inferred source cell type, and a target cell type where that factor
is called as potential admixture. We only print the first few columns and rows to
avoid dumping a large table.

Each rule also passes a native-factor false-positive check: target cells with
no source-type cells among their nearest neighbors are examined, and rules
whose factor persists in those source-distant cells are flagged
(`keep = FALSE`, with the reason in `native_check`) and skipped by
correction. This guards against removing factors that are genuinely native to
a related target type — on this dataset, the malignant-cell factor in
epithelial cells is the canonical case: epithelial cells far from any
malignant cell still express it, so it should not be removed from them.

```{r bridge-rules}
rules <- bridge$rules()
rules[, c("factor", "source_cell_type", "target_cell_type", "p_value",
  "keep", "native_check", "native_distant_median")]
```

The heatmap shows which source cell types are implicated for each factor and
target cell type. It is a compact way to check whether the factor annotations
make biological sense before applying correction.

```{r bridge-heatmap, fig.width=8, fig.height=5}
bridge$plot_heatmap()
```

The pair plots show the same bridge evidence one factor at a time. They make it
easier to inspect whether inferred source cell types and significant target cell
types form a coherent source-target pattern.

```{r bridge-score-pairs, fig.width=score_pairs_width, fig.height=score_pairs_height}
bridge$plot_pairs()
```

## Cleanup

Cleanup removes molecules matching the bridge-supported source-target-factor
rules. The operation writes a corrected run into the output directory and keeps
the original fit unchanged.

```{r apply-correction}
correction <- bridge$correct(rules = rules)

correction$summary()
```

The next plot summarizes how many molecules were removed in modified cells.
Large removals should be inspected carefully, while very small removals may have
little impact on downstream summaries.

```{r molecule-removal, fig.width=6.5, fig.height=5}
correction$plot_removed_molecules()
```

## Correction Diagnostics

For this example, the diagnostic question is whether malignant epithelial
marker signal in fibroblast cells is reduced after cleanup, while native
fibroblast marker signal is preserved. These plots are not required for
cellAdmix itself; they are an interpretable downstream check for this dataset.

```{r diagnostic-data}
malignant_marker_genes <- c(
  "KRT19", "KRT8", "KRT18", "TACSTD2", "CEACAM6", "CDH1", "EPCAM",
  "CLDN4", "SFN", "PHLDA2", "DDR1", "AGR2", "KRT6A", "KRT17",
  "CD24", "KRT5", "AQP3", "SLC2A1", "KRT7", "COL17A1", "FASN",
  "KRT15", "ITGA2", "ERBB3", "S100A2", "SPINK1", "ITGB4", "PIGR",
  "SOX2", "ITGA3", "CYSTM1", "NTRK2", "MAPK13", "CCND1", "HDAC1", "EGFR"
)
malignant_label_genes <- c("KRT19", "KRT8", "KRT17", "CEACAM6")
fibroblast_marker_genes <- c("MYL9", "CXCL12", "PDGFRB", "RARRES2")

counts <- fit$counts()
fib_cells <- rownames(cell_meta)[
  cell_meta$celltype == "fibroblast" & !is.na(cell_meta$regions_compare)
]
fib_cells <- intersect(fib_cells, colnames(counts))
regions_compare <- setNames(cell_meta[fib_cells, "regions_compare"], fib_cells)
de_out_orig <- celladmix_de(counts[, fib_cells, drop = FALSE], regions_compare,
  contrast = c("tumor", "stroma"))
impact <- celladmix_correction_impact(correction, counts, fib_cells,
  regions_compare, contrast = c("tumor", "stroma"))

data.frame(fibroblast_cells = length(fib_cells),
  regions = paste(sort(unique(na.omit(regions_compare))), collapse = ", "))
```

The uncorrected volcano plot compares fibroblast cells in tumor-associated
regions with fibroblast cells in stromal regions. Malignant markers enriched in
the tumor-associated fibroblasts are the signal we expect cleanup to reduce.

```{r baseline-volcano, fig.width=6, fig.height=4}
celladmix_plot_volcano(de_out_orig, markers = malignant_marker_genes,
  label = malignant_label_genes, marker_label = "Malignant marker",
  title = "Before cleanup")
```

The marker barplots compare mean normalized expression before and after cleanup
within fibroblast cells. The desired pattern is preservation of native
fibroblast markers and reduction of malignant marker signal.

```{r marker-expression, fig.width=8, fig.height=3.8}
celladmix_plot_marker_expression(counts, impact$counts, fib_cells,
  markers = list("Fibroblast markers" = fibroblast_marker_genes,
    "Malignant markers" = malignant_label_genes))
```

The cleaned volcano plot repeats the same fibroblast comparison after molecule
removal. A successful correction should shift suspicious malignant markers
toward weaker significance and smaller tumor-region enrichment.

```{r cleaned-volcano, fig.width=6, fig.height=4}
celladmix_plot_volcano(impact$de, markers = malignant_marker_genes,
  label = malignant_label_genes, marker_label = "Malignant marker",
  title = "After cleanup")
```

The final scatter plot directly compares signed adjusted-p-value significance
before and after cleanup. Positive values indicate tumor-region enrichment
within fibroblasts, negative values indicate stromal-region enrichment, and
points below the diagonal became less significant after cleanup. Highlighted
malignant markers moving toward zero are consistent with reduced malignant
admixture signal.

```{r pvalue-scatter, fig.width=5.5, fig.height=4.6}
celladmix_plot_de_shift(de_out_orig, impact$de,
  markers = malignant_marker_genes, label = malignant_label_genes,
  marker_label = "Malignant marker")
```

The expected result is not that every malignant marker disappears. Instead, the
correction should remove the molecule subset assigned to bridge-supported
admixture factors, reduce the strongest suspicious marker signal in target
cells, and leave native fibroblast markers largely intact.
