---
title: "Generative Admixture Correction on CosMx NSCLC"
output:
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---

```{r setup, include=FALSE}
library(cellAdmixCore)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE,
  fig.align = "center", dpi = 96, fig.retina = 2)
```

The generative model corrects admixture by decomposing each cell's counts
into the cell's own expression, contamination from each neighboring cell
type, ambient background, and expression the cell switched on in response
to its neighborhood; the contamination and ambient shares are removed,
the induced expression retained
([docs/generative.md](../../docs/generative.md)). This page runs it on
the CosMx NSCLC dataset and characterizes both sides of the separation:
the pattern of material moving between cell types, and the induced
programs the correction preserves. One platform note applies throughout:
this CosMx export carries no nucleus-overlap flag, so the model uses
whole-cell rather than cytoplasmic source profiles.

```{r fit}
cell_meta <- read.csv(file.path("prepared", "cell_metadata_all.csv.gz"),
  stringsAsFactors = FALSE)
cell_annotation <- setNames(cell_meta$cell_type_coarse, cell_meta$cell)

ds <- cellAdmix(file.path("prepared", "molecules_all.csv.gz"),
  output_dir = "out", annotation = cell_annotation)
nmf_fit <- ds$fit()
```

## The admixture pattern between cell types

```{r audit-map, fig.width=6.5, fig.height=5}
audit <- nmf_fit$audit_admixture()
audit$plot_map()
```

```{r audit-pairs}
pairs <- audit$pairs(detected_only = TRUE)
head(pairs[order(-pairs$rate),
  c("source", "target", "admixed_molecules", "rate")], 8)
```

Aggregated over targets, the shedding is dominated by the tumor and
myeloid compartments:

```{r audit-sources, fig.width=6, fig.height=3}
by_source <- aggregate(admixed_molecules ~ source, pairs, sum)
ggplot(by_source, aes(reorder(source, admixed_molecules),
    admixed_molecules / 1e3)) +
  geom_col(fill = "#c0392b", alpha = 0.85) +
  coord_flip() +
  labs(x = NULL, y = "estimated admixed molecules shed (thousands)") +
  theme_classic(base_size = 11)
```

## Correcting with the generative model

The model is fitted on the audit's detected pairs (the NMF fit only
initializes its expression programs), and the correction derives from
the fitted model:

```{r correct}
model <- audit$fit_generative(init = nmf_fit, num_threads = 8)
model
correction <- model$correct()
```

The per-cell contamination fractions for the largest pair, by exposure
and in space:

```{r composition, fig.width=6, fig.height=3.6}
top_pair <- pairs[order(-pairs$admixed_molecules), ][1, ]
comp <- model$composition(top_pair$source, top_pair$target)
cells_xy <- nmf_fit$cell_factors()
expo <- setNames(
  cellAdmixCore:::.celladmix_source_exposure_counts(
    cells_xy, cell_annotation, 15L)$counts[, top_pair$source],
  as.character(cells_xy$cell_id))
comp$exposure <- pmin(expo[comp$cell_id], 4)
ggplot(comp, aes(factor(exposure), contamination)) +
  geom_boxplot(outlier.size = 0.3, fill = "#f4a582") +
  labs(x = sprintf("%s cells among 15 nearest neighbors", top_pair$source),
    y = sprintf("fraction of the %s cell's molecules\nattributed to %s contamination",
      top_pair$target, top_pair$source)) +
  theme_classic(base_size = 11)
```

```{r composition-map, fig.width=7.5, fig.height=5}
xy <- cells_xy[match(comp$cell_id, as.character(cells_xy$cell_id)), ]
ggplot(cbind(comp, x = xy$x, y = xy$y), aes(x, y,
    color = pmin(contamination, 0.5))) +
  geom_point(size = 0.2) +
  scale_color_viridis_c(option = "inferno",
    name = sprintf("%s share\nof %s content",
      top_pair$source, top_pair$target)) +
  coord_equal() +
  labs(x = NULL, y = NULL) +
  theme_void(base_size = 11)
```

## The induction pattern

On this tissue the retained induced set reads as the classic stress,
immediate-early and chemokine programs of cells at tissue interfaces:

```{r induced}
induced <- model$induced
head(induced[order(-induced$excess),
  c("source", "target", "gene", "excess", "fold", "z")], 10)
```

The contrast between a transferred and an induced gene, for the pair with
the strongest induced signal — the observed rise of the transferred gene
is flattened to the unexposed level, while the induced gene keeps its
gradient:

```{r gene-contrast, fig.width=8.5, fig.height=3.6}
ind_top <- induced[order(-induced$excess), ][1, ]
mk <- audit$markers(ind_top$source, ind_top$target)
transfer_gene <- setdiff(mk$pool, induced$gene)[1]
before <- nmf_fit$counts()
after <- correction$counts()
t_cells <- model$composition(ind_top$source, ind_top$target)$cell_id
expo_t <- setNames(
  cellAdmixCore:::.celladmix_source_exposure_counts(
    cells_xy, cell_annotation, 15L)$counts[, ind_top$source],
  as.character(cells_xy$cell_id))[t_cells]
tot <- Matrix::colSums(before[, t_cells])
bins <- pmin(expo_t, 3)
rate_by_bin <- function(m, gene) {
  tapply(m[gene, t_cells], bins, sum) / tapply(tot, bins, sum) * 1e3
}
df <- do.call(rbind, lapply(c(transfer_gene, ind_top$gene), function(g) {
  rbind(
    data.frame(gene = g, counts = "observed", bin = 0:3,
      rate = as.numeric(rate_by_bin(before, g))),
    data.frame(gene = g, counts = "corrected", bin = 0:3,
      rate = as.numeric(rate_by_bin(after, g))))
}))
df$gene <- factor(df$gene, levels = c(transfer_gene, ind_top$gene),
  labels = c(sprintf("%s (transferred)", transfer_gene),
    sprintf("%s (induced)", ind_top$gene)))
ggplot(df, aes(bin, rate, color = counts)) +
  geom_line() + geom_point() +
  facet_wrap(~gene, scales = "free_y") +
  scale_color_manual(values = c(observed = "#c0392b",
    corrected = "#2980b9")) +
  labs(x = sprintf("%s cells among 15 nearest neighbors", ind_top$source),
    y = sprintf("rate in %s cells (per 1,000 molecules)", ind_top$target)) +
  theme_classic(base_size = 11)
```

## Verification

```{r verify, fig.width=8, fig.height=4}
report <- audit$evaluate(correction)
report$summary()
report$plot_cleanup()
```

The model never removes a target type's own marker genes, and induced
expression is only recognizable when disproportionate to the source's
transfer profile — see the identifiability limit in
[docs/generative.md](../../docs/generative.md).
