---
title: "Generative Admixture Correction on Xenium Pancreas"
output:
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---

```{r setup, include=FALSE}
library(cellAdmixCore)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE,
  fig.align = "center", dpi = 96, fig.retina = 2)
```

Molecules leak between neighboring cells, so a cell's counts mix its own
expression with material from the cells around it. But not every count
that rises near a neighbor is leakage: cells also change their own
transcription in response to their neighborhood, and a correction that
flattens every neighbor-linked gradient erases exactly that biology. The
generative model separates the two: it decomposes each cell's counts into
the cell's own expression programs, contamination from each neighboring
cell type, ambient background, and neighborhood-induced expression, then
removes the contamination and ambient shares while keeping the induced
ones ([docs/generative.md](../../docs/generative.md) describes the model
and its validation; [the minimal version](pancreas_generative_minimal.md)
of this page is four calls long).

This example characterizes both sides on the pancreas dataset: how much
material moves between which cell types, and which genes the model
recognizes as the cells' own induced expression.

```{r fit}
annotation <- read.csv(file.path("annotations", "annotation.csv.gz"))
cell_annotation <- setNames(annotation$merged_annotation, annotation$cell_id)

ds <- cellAdmix("data", output_dir = "out", annotation = cell_annotation)
nmf_fit <- ds$fit(nmf_variant = "invsqrt_kl")
```

## The admixture pattern between cell types

The audit measures, for every ordered pair of cell types, how many
molecules leaked from the source type into cells of the target type,
using the rise of source-marker content with the number of source-type
neighbors:

```{r audit-map, fig.width=6.5, fig.height=5}
audit <- nmf_fit$audit_admixture()
audit$plot_map()
```

The magnitudes are far from uniform. Sorting the detected pairs by their
estimated admixture rate — the fraction of the target type's molecules
that originated in the source:

```{r audit-pairs}
pairs <- audit$pairs(detected_only = TRUE)
head(pairs[order(-pairs$rate),
  c("source", "target", "admixed_molecules", "rate")], 8)
```

Aggregating over targets shows which cell types shed the most material —
on this tissue, the exocrine compartment dominates as a source:

```{r audit-sources, fig.width=6, fig.height=3.2}
by_source <- aggregate(admixed_molecules ~ source, pairs, sum)
ggplot(by_source, aes(reorder(source, admixed_molecules),
    admixed_molecules / 1e3)) +
  geom_col(fill = "#c0392b", alpha = 0.85) +
  coord_flip() +
  labs(x = NULL, y = "estimated admixed molecules shed (thousands)") +
  theme_classic(base_size = 11)
```

## Correcting with the generative model

The model is fitted on the audit's detected pairs; the NMF fit is passed
only to initialize the model's expression programs from its
factor-labeled molecules (`init = "clusters"` or `init = "pseudobulk"`
fit without an NMF fit, with nearly identical results). The correction
then derives from the fitted model:

```{r correct}
model <- audit$fit_generative(init = nmf_fit, num_threads = 8)
model
correction <- model$correct()
before <- nmf_fit$counts()
after <- correction$counts()
```

The most quotable summary of the result is the admixture burden per cell
type: the fraction of each type's total molecule content that the model
attributed to contamination or ambient background and removed.

```{r burden, fig.width=6, fig.height=3.2}
cell_types <- cell_annotation[colnames(before)]
removed_by_type <- tapply(Matrix::colSums(before) - Matrix::colSums(after),
  cell_types, sum)
total_by_type <- tapply(Matrix::colSums(before), cell_types, sum)
burden <- data.frame(type = names(removed_by_type),
  share = as.numeric(removed_by_type / total_by_type))
ggplot(burden, aes(reorder(type, share), share)) +
  geom_col(fill = "#c0392b", alpha = 0.85) +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(x = NULL, y = "share of the type's molecules removed as admixture") +
  theme_classic(base_size = 11)
```

The model's central output is the per-cell decomposition. Averaged over
ductal cells at each exocrine exposure, it shows where a contaminated
cell's molecules come from: the cell's own expression share falls with
exposure as the exocrine contamination share grows, with smaller
contributions from the other bordering types and the ambient background:

```{r composition, fig.width=7, fig.height=3.8}
comp_all <- model$composition(target = "Ductal/tumor epithelial")
cells_xy <- nmf_fit$cell_factors()
expo <- setNames(
  cellAdmixCore:::.celladmix_source_exposure_counts(
    cells_xy, cell_annotation, 15L)$counts[, "Exocrine epithelial"],
  as.character(cells_xy$cell_id))
comp_all$bin <- factor(pmin(expo[comp_all$cell_id], 3), levels = 0:3,
  labels = c("0", "1", "2", "3+"))
shares <- aggregate(contamination ~ bin + source, comp_all, mean)
ambient <- aggregate(ambient ~ bin, comp_all[!duplicated(comp_all$cell_id), ],
  mean)
shares <- rbind(shares,
  data.frame(bin = ambient$bin, source = "ambient",
    contamination = ambient$ambient))
own <- aggregate(contamination ~ bin, shares, sum)
shares <- rbind(shares,
  data.frame(bin = own$bin, source = "own expression",
    contamination = 1 - own$contamination))
shares$source <- factor(shares$source,
  levels = c("own expression", "ambient",
    rev(unique(comp_all$source))))
ggplot(shares, aes(bin, contamination, fill = source)) +
  geom_col(width = 0.72) +
  scale_fill_manual(values = c("own expression" = "#a8c6df",
    "ambient" = "#8a8a8a",
    setNames(hcl.colors(length(unique(comp_all$source)), "Reds 3"),
      rev(unique(comp_all$source))))) +
  labs(x = "exocrine cells among 15 nearest neighbors",
    y = "mean share of ductal-cell content", fill = NULL) +
  theme_classic(base_size = 11)
```

The same fractions drawn in space show the admixture concentrating at
the tissue interfaces: ductal cells inside the acinar tissue carry up to
half exocrine content, while the ductal/tumor mass itself is nearly
clean:

```{r composition-map, fig.width=7, fig.height=5}
comp <- model$composition("Exocrine epithelial",
  "Ductal/tumor epithelial")
xy <- cells_xy[match(comp$cell_id, as.character(cells_xy$cell_id)), ]
ggplot(cbind(comp, x = xy$x, y = xy$y), aes(x, y,
    color = pmin(contamination, 0.6))) +
  geom_point(size = 0.25) +
  scale_color_viridis_c(option = "inferno",
    name = "exocrine share\nof ductal-cell\ncontent") +
  coord_equal() +
  labs(x = NULL, y = NULL,
    title = "Exocrine contamination of ductal cells") +
  theme_void(base_size = 11)
```

The transferred material itself is profile-shaped, not uniform: the
genes removed from ductal cells are dominated by the source's largest
expression channels, the digestive enzymes of the exocrine pancreas:

```{r removed-genes}
duct_cells <- comp$cell_id
removed_g <- Matrix::rowSums(before[, duct_cells]) -
  Matrix::rowSums(after[, duct_cells])
head(data.frame(gene = names(sort(removed_g, decreasing = TRUE)),
  removed = round(sort(removed_g, decreasing = TRUE))), 8)
```

## The induction pattern

The model retains genes whose exposure-linked excess exceeds what
transfer from the source can deliver — expression the target cells
switched on themselves. On this tissue the retained set is led by the
duct-cell program that ductal cells activate at acinar interfaces:

```{r induced}
induced <- model$induced
head(induced[order(-induced$excess),
  c("source", "target", "gene", "excess", "fold", "z")], 8)
```

The evidence behind these calls is a proportionality test. Transferred
material samples the source's transcriptome, so a gene's excess in
ductal cells should be proportional to its share of the exocrine
profile; genes far above that expectation are the ductal cells' own
response. Each point below is one exocrine-owned gene, with the genes
the model flagged in this pair highlighted:

```{r screen, fig.width=6.2, fig.height=4.6}
exp_cells <- duct_cells[expo[duct_cells] > 0]
un_cells <- duct_cells[expo[duct_cells] == 0]
t_exp <- sum(Matrix::colSums(before[, exp_cells]))
t_un <- sum(Matrix::colSums(before[, un_cells]))
profiles <- cellAdmixCore:::.celladmix_audit_profiles(before, cell_types)
gset <- rownames(profiles)[
  colnames(profiles)[max.col(profiles, ties.method = "first")] ==
    "Exocrine epithelial"]
S_cells <- names(cell_types)[!is.na(cell_types) &
  cell_types == "Exocrine epithelial"]
dist_T <- cellAdmixCore:::.celladmix_source_nearest_distance(
  cells_xy, cell_annotation)[, "Ductal/tumor epithelial"]
psi <- cellAdmixCore:::.celladmix_audit_interface_profile(
  before, S_cells, dist_T)[gset]
m_exp <- Matrix::rowSums(before[gset, exp_cells])
m_un <- Matrix::rowSums(before[gset, un_cells])
excess <- m_exp - m_un / t_un * t_exp
v <- m_exp + (t_exp / t_un)^2 * m_un + 1
slope <- max(sum(excess * psi / v) / sum(psi^2 / v), 0)
scr <- data.frame(gene = gset, excess = excess, expected = slope * psi)
scr$flagged <- scr$gene %in%
  induced$gene[induced$source == "Exocrine epithelial" &
    induced$target == "Ductal/tumor epithelial"]
scr <- scr[scr$excess > 10 & scr$expected > 1, ]
ggplot(scr, aes(expected, excess)) +
  geom_abline(color = "grey40", linewidth = 0.4) +
  geom_point(data = scr[!scr$flagged, ], size = 1.3, color = "#9bb5c9") +
  geom_point(data = scr[scr$flagged, ], size = 2.2, shape = 15,
    color = "#c0392b") +
  ggrepel::geom_text_repel(
    data = scr[scr$flagged | scr$excess > 5e4, ],
    aes(label = gene), size = 3) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "excess expected from proportional exocrine transfer (molecules)",
    y = "measured excess in ductal cells (molecules)") +
  theme_classic(base_size = 11)
```

The consequence is visible gene by gene. AMY2A sits on the diagonal
above — the largest exocrine transfer channel — and the correction
flattens its exposure gradient to the unexposed level. CFTR sits far
above the diagonal, and the correction preserves its gradient, removing
only the transferred share. The same fitted model can also derive the
opposite policy — `model$correct(retain_induced = FALSE)` removes the
fitted induced share along with the contamination — which shows directly
what retention protects, at no change on the transferred gene:

```{r gene-contrast, fig.width=8.5, fig.height=3.6}
correction_noind <- model$correct(name = "generative_noind",
  retain_induced = FALSE)
after_noind <- correction_noind$counts()
tot <- Matrix::colSums(before[, duct_cells])
bins <- pmin(expo[duct_cells], 3)
rate_by_bin <- function(m, gene) {
  tapply(m[gene, duct_cells], bins, sum) / tapply(tot, bins, sum) * 1e3
}
df <- do.call(rbind, lapply(c("AMY2A", "CFTR"), function(g) {
  do.call(rbind, lapply(list(
    list(m = before, lab = "observed"),
    list(m = after, lab = "corrected"),
    list(m = after_noind, lab = "corrected, induced share removed")), function(a) {
    data.frame(gene = g, counts = a$lab, bin = 0:3,
      rate = as.numeric(rate_by_bin(a$m, g)))
  }))
}))
df$counts <- factor(df$counts,
  levels = c("observed", "corrected", "corrected, induced share removed"))
ggplot(df, aes(bin, rate, color = counts, linetype = counts)) +
  geom_line() + geom_point(size = 1.6) +
  facet_wrap(~gene, scales = "free_y") +
  scale_color_manual(values = c("observed" = "#c0392b",
    "corrected" = "#2980b9", "corrected, induced share removed" = "#7d3c98")) +
  scale_linetype_manual(values = c("observed" = "solid",
    "corrected" = "solid", "corrected, induced share removed" = "22")) +
  labs(x = "exocrine cells among 15 nearest neighbors",
    y = "rate in ductal cells (per 1,000 molecules)") +
  theme_classic(base_size = 11)
```

Across all retained genes, most of the exposure-linked excess survives
the correction — the retained fraction per gene, for the largest calls:

```{r retention, fig.width=7.5, fig.height=3.6}
gene_excess <- function(m, gene, source, target) {
  t_cells <- names(cell_types)[!is.na(cell_types) & cell_types == target]
  e <- setNames(cellAdmixCore:::.celladmix_source_exposure_counts(
    cells_xy, cell_annotation, 15L)$counts[, source],
    as.character(cells_xy$cell_id))[t_cells]
  tt <- Matrix::colSums(before[, t_cells])
  sum(m[gene, t_cells[e > 0]]) -
    sum(m[gene, t_cells[e == 0]]) / max(sum(tt[e == 0]), 1) * sum(tt[e > 0])
}
top_ind <- head(induced[order(-induced$excess), ], 8)
ret <- do.call(rbind, lapply(seq_len(nrow(top_ind)), function(i) {
  r <- top_ind[i, ]
  data.frame(gene = r$gene,
    counts = c("observed excess", "excess after correction"),
    excess = c(gene_excess(before, r$gene, r$source, r$target),
      gene_excess(after, r$gene, r$source, r$target)))
}))
ret$counts <- factor(ret$counts,
  levels = c("observed excess", "excess after correction"))
ggplot(ret, aes(reorder(gene, -excess), excess / 1e3, fill = counts)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("observed excess" = "#c0392b",
    "excess after correction" = "#1e8449")) +
  labs(x = NULL, y = "exposure-linked excess (thousand molecules)",
    fill = NULL) +
  theme_classic(base_size = 11)
```

The retained program has a spatial identity of its own. Drawing the
corrected CFTR expression across ductal cells shows it peaking exactly
where the induced table says it should — the duct cells embedded in
acinar tissue — a pattern a marker-based correction would have removed
as contamination:

```{r cftr-map, fig.width=7, fig.height=5}
cftr_rate <- after["CFTR", duct_cells] /
  pmax(Matrix::colSums(after[, duct_cells]), 1) * 1e3
ggplot(cbind(comp, x = xy$x, y = xy$y, cftr = pmin(cftr_rate, 250)),
    aes(x, y, color = cftr)) +
  geom_point(size = 0.25) +
  scale_color_viridis_c(name = "corrected CFTR\n(per 1,000\nmolecules)") +
  coord_equal() +
  labs(x = NULL, y = NULL,
    title = "The retained CFTR program in ductal cells") +
  theme_void(base_size = 11)
```

## Verification

The correction is verified against the audit's own measurements, like
any other correction in the package. The pooled exposure profile shows
the flattening directly:

```{r verify, fig.width=8, fig.height=4}
report <- audit$evaluate(correction)
report$summary()
report$plot_cleanup()
```

```{r verify-exposure, fig.width=6, fig.height=4}
audit$plot_exposure(correction = correction)
```

Two limits are worth keeping in mind when reading the results. The model
never removes a target type's own marker genes (their contamination is
indistinguishable from native expression and is deliberately left in
place), so shared-gene admixture is under-removed by construction. And
induced expression is only recognizable when it is disproportionate to
the source's transfer profile: induction of a source's own top marker
genes cannot be told apart from admixture at count level and is removed
with it ([docs/generative.md](../../docs/generative.md), the
identifiability limit).
