---
title: "cellAdmix Xenium Example: Pancreas Membrane and Bridge Scoring"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
library(cellAdmixCore)
library(ggplot2)
library(cowplot)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE,
  fig.align = "center", out.width = "100%", dpi = 150)
par(mar = c(3.5, 3.5, 2.0, 0.5), mgp = c(2, 0.65, 0), cex = 2/3)
```

## Overview

This notebook illustrates a cellAdmix workflow on a Xenium pancreas dataset
with membrane staining. The example starts from a Xenium output bundle, a simple
cell-type annotation, and a coarse regional annotation. The core fit uses
default cellAdmix settings: inverse-square-root KL NMF, automatic rank from the
number of annotated cell types, multiple NMF starts, and gene-loading molecule
scoring.

At a high level, the workflow is:

```{r api-overview, eval=FALSE}
ds <- cellAdmix(bundle_dir, output_dir = output_dir, annotation = cell_annotation)
fit <- ds$fit()

membrane_score <- fit$score_membrane()
membrane_rules <- membrane_score$rules()
membrane_correction <- membrane_score$correct(rules = membrane_rules)

bridge_score <- fit$score_bridge()
bridge_rules <- bridge_score$rules()
bridge_correction <- bridge_score$correct(rules = bridge_rules)
```

The constructor records the Xenium bundle location, validates the input
configuration, and stores lightweight cell-level metadata. It does not load all
molecules into R memory. The computational work starts with `fit()`, followed by
one or more scoring methods that identify source-target admixture patterns and
produce correction rules.

## Inputs

Download the Xenium output bundle from the original
[10x Genomics dataset page](https://www.10xgenomics.com/datasets/ffpe-human-pancreas-with-xenium-multimodal-cell-segmentation-1-standard)
and expand it into a local folder named `data`. The bundle zip can be
downloaded directly with:

```{r download-bundle-command, eval=FALSE}
curl -O https://cf.10xgenomics.com/samples/xenium/2.0.0/Xenium_V1_human_Pancreas_FFPE/Xenium_V1_human_Pancreas_FFPE_outs.zip
```

Also create a folder named `annotations` and download the example annotation
files from
<http://pklab.org/peterk/cellAdmix/examples/pancreas_377/annotations>.
The two files used below are `annotations/annotation.csv.gz` and
`annotations/domain_annotation.csv.gz`. Base R can read these gzip-compressed
CSV files directly.

The supplied cell annotation and regional-domain segmentation files are
quick-and-dirty labels generated only for this illustration. They are not
curated biological ground truth, and they are probably wrong in places. In a
real analysis, use your own cell-type annotations and tissue-region definitions.

The output directory `out` stores reusable cellAdmix caches, fit outputs,
scores, and corrected runs.

```{r inputs}
bundle_dir <- "data"
output_dir <- "out"
annotation_file <- file.path("annotations", "annotation.csv.gz")
domain_file <- file.path("annotations", "domain_annotation.csv.gz")

annotation <- read.csv(annotation_file, stringsAsFactors = FALSE)
domain_annotation <- read.csv(domain_file, stringsAsFactors = FALSE)

table(annotation$merged_annotation)
table(domain_annotation$domain_label)
```

The cell-type annotation is passed to `cellAdmix()` as a named vector. The
regional-domain table is kept in R memory because it is only used later for
plots and diagnostic DE comparisons, not for factor fitting or scoring.

```{r construct-celladmix}
cell_annotation <- setNames(annotation$merged_annotation, annotation$cell_id)

ds <- cellAdmix(bundle_dir, output_dir = output_dir, annotation = cell_annotation)

ds
```

## Fit Factors

`ds$fit()` constructs molecule-centered neighborhood composition vectors, fits
NMF factors, projects/smooths molecule factor assignments, and summarizes the
result at cell level. The rank is automatically derived from the number of cell
types in the active annotation.

```{r fit-default}
fit <- ds$fit(verbose = TRUE)

fit
```

Top gene loadings are the first-pass interpretation of each factor.

```{r factor-loadings, fig.width=8.5, fig.height=5.0}
fit$plot_loadings()
```

The stability plot shows whether high-importance factors are reproducible
across independent NMF restarts (matched gene-ownership correlation: near
zero for unrelated factors, above 0.3 for reproducibly re-found ones).

```{r factor-stability, fig.width=6.2, fig.height=4.2}
fit$plot_stability()
```

## Factor Source Prior

Top loadings are useful, but source calls should emphasize genes that actually
distinguish annotated cell types. `score_factor_sources()` builds pseudobulk
cell-type profiles from the original counts, estimates one-versus-rest
cell-type marker weights, and compares each NMF factor against those weighted
marker signatures. This is a diagnostic source prior only; the membrane and
bridge scores below are still run independently.

```{r factor-source-score}
original_counts <- fit$counts()
factor_source_score <- fit$score_factor_sources(counts = original_counts)

factor_source_score$annotation()
```

In this table, `score` is the marker-weighted similarity between a factor and
the best matching cell type. `margin` is the best score minus the second-best
score for that factor; larger margins indicate more specific source evidence.
`called` indicates whether the best cell type passed both the score and margin
thresholds used for source-prior annotation.

The heatmap shows marker-weighted gene-content evidence for each factor and
cell type. The `S` marker identifies the best gene-content source call for each
factor.

```{r factor-source-heatmap, fig.width=8.5, fig.height=5.0}
factor_source_score$plot_heatmap()
```

The next plot shows the cell-type marker genes that most strongly support each
factor's best source call.

```{r factor-source-top-genes, fig.width=9.0, fig.height=5.8}
factor_source_score$plot_top_genes()
```

## Tissue Context

The cell-type and domain plots below are side visualizations. To prepare them,
we merge the cellAdmix cell-level factor summaries with the supplied cell-type
and regional-domain annotations.

```{r cell-spatial-data}
cell_factors <- fit$cell_factors()
cell_spatial <- merge(cell_factors,
  annotation[, c("cell_id", "merged_annotation", "cluster_label")],
  by = "cell_id", all.x = FALSE, sort = FALSE)
cell_spatial <- merge(cell_spatial,
  domain_annotation[, c("cell_id", "domain_label", "domain_fine_label")],
  by = "cell_id", all.x = FALSE, sort = FALSE)
cell_spatial$domain_label <- factor(cell_spatial$domain_label,
  levels = c("adjacent_normal", "transition", "tumor"))
```

```{r spatial-celltypes, fig.width=6.0, fig.height=4.8}
celladmix_plot_spatial(cell_spatial, color_by = "merged_annotation") +
  ggtitle("Cell type annotation")
```

```{r spatial-domains-factors, fig.width=11.0, fig.height=4.8}
cowplot::plot_grid(plotlist = list(
  celladmix_plot_spatial(cell_spatial, color_by = "domain_label") +
    ggtitle("Coarse tissue domains"),
  celladmix_plot_spatial(cell_spatial, color_by = "dominant_factor") +
    ggtitle("Dominant cellAdmix factor")
), ncol = 2, align = "hv", axis = "tblr")
```

## Diagnostic Setup

Scoring and correction below are run for all annotated cell types. To keep the
diagnostic plots readable, we focus the downstream DE and example-cell panels on
two illustrative cell types: endothelial and immune cells. These choices are
only for visualization; they do not restrict the cleanup rules.

The marker genes below are used to highlight likely admixture sources in the
diagnostic plots: ductal/tumor epithelial cells and endocrine/exocrine
epithelium.

```{r diagnostic-marker-sets}
diagnostic_cell_types <- c("Endothelial", "Immune")

ductal_tumor_markers <- c("GPRC5A", "MALL", "FHL2", "EPCAM", "GPX2", "MET",
  "SERPINB3", "KRT7")
endocrine_exocrine_markers <- c("AMY2A", "AQP8", "ANPEP", "CFTR", "INS",
  "GCG", "SST", "CHGA", "SCGN")
source_marker_sets <- list(
  "ductal/tumor source" = ductal_tumor_markers,
  "endocrine/exocrine source" = endocrine_exocrine_markers
)
```

We then run a baseline DE comparison between tumor and adjacent-normal domains
for the two diagnostic cell types using the original counts collected above.
These DE results are only used for the illustrative before/after cleanup plots
later in the notebook.

```{r diagnostic-de-original}
diagnostic_de_original <- celladmix_de_by_group(original_counts, cell_spatial,
  subset_by = "merged_annotation", subsets = diagnostic_cell_types,
  group_by = "domain_label", contrast = c("tumor", "adjacent_normal"))
```

These baseline volcano plots show tumor-domain versus adjacent-normal-domain
differences before any molecule removal. Positive logFC means higher expression
in tumor-domain target cells.

```{r baseline-volcano, fig.width=11.0, fig.height=4.8}
baseline_volcano_plots <- lapply(diagnostic_cell_types, function(target_type) {
  de <- diagnostic_de_original[[target_type]]
  celladmix_plot_volcano(de, markers = source_marker_sets,
    title = paste(target_type, "before cleanup"),
    subtitle = sprintf("logFC = tumor / adjacent_normal; n = %d / %d",
      unique(de$n_group_a), unique(de$n_group_b)))
})
cowplot::plot_grid(plotlist = baseline_volcano_plots,
  ncol = length(baseline_volcano_plots), align = "hv", axis = "tblr")
```

## Membrane Scoring

Membrane scoring asks whether target-cell molecules assigned to a factor are
enriched along membrane-stain paths facing a candidate source cell type.
The score summaries are thresholded at `p_thresh`; here we use a permissive
threshold because this notebook is meant to compare candidate cleanup rules.

```{r score-settings}
p_thresh <- 0.1
score_pairs_width <- 9.6
score_pairs_height <- max(6, ceiling(fit$rank / 2) * 4.9)
```

First, run the membrane scoring procedure. The returned `membrane_score` object
stores the raw pair-level and summarized evidence; it does not modify molecules.

```{r membrane-score}
membrane_score <- fit$score_membrane()
```

Next, convert score summaries into factor annotations. For each factor, this
identifies the most plausible source cell type and target cell types with
significant membrane-supported admixture evidence.

```{r membrane-annotation}
membrane_annotation <- membrane_score$annotation(p_thresh = p_thresh)
```

Finally, convert those annotation calls into correction rules. Each row says that
molecules assigned to a given factor should be removed from a target cell type;
the source cell type is retained as an interpretation of the likely origin.

```{r membrane-rules}
membrane_rules <- membrane_score$rules(p_thresh = p_thresh)

head(membrane_rules[, c("factor", "source_cell_type", "target_cell_type", "p_value")])
```

The heatmap summarizes factor-by-cell-type evidence. Color intensity shows the
strongest membrane-supported target evidence for each factor and cell type; plot
symbols mark inferred source and cleanup target calls. With the gene-content
source prior overlaid, `S` marks the membrane/spatial source, `G` marks the
gene-content source, `SG` marks agreement, separate `S` and `G` labels in the
same factor column indicate a source conflict, and `*` marks a cleanup target.

```{r membrane-heatmap, fig.width=8.5, fig.height=5.4}
membrane_score$plot_heatmap(p_thresh = p_thresh,
  source_prior = factor_source_score)
```

The pair plots show the same membrane evidence one factor at a time. Rows are
factors, columns are candidate target cell types, and the top margin summarizes
which source cell types best explain each factor. This view is useful for
checking whether a correction rule has a plausible source and target pattern.

```{r membrane-score-pairs, fig.width=score_pairs_width, fig.height=score_pairs_height}
membrane_score$plot_pairs()
```

### Example Cells

Next, we render a few target-cell examples for the diagnostic cell types. The
cell selection uses the membrane score to find cells with strong rule-supported
admixture-factor evidence. The molecule colors show post-smoothing factor
labels: native factors in blue, the top putative admixture factor in red, and
other non-native factors in orange. Marker-gene molecules are shown with the
same color coding, but are drawn slightly larger.

The DAPI/membrane stain backgrounds, cell boundaries, and cell-type shading
are discovered automatically from the Xenium bundle; stain pixel data are read
lazily as small crops around each selected cell. Pass `stains = NULL` or
`shade_cell_types = FALSE` to simplify the panels, or plot specific cells with
`plot_examples(cells = c(...))`.

```{r membrane-example-overlays, fig.width=10.0, fig.height=10.0}
example_cells <- membrane_score$examples(rules = membrane_rules,
  score_annotation = membrane_annotation, cell_data = cell_spatial,
  targets = diagnostic_cell_types, n_per_target = 4)

for (cell_type in diagnostic_cell_types) {
  type_examples <- example_cells[example_cells$target_cell_type == cell_type, , drop = FALSE]
  print(membrane_score$plot_examples(type_examples,
    score_annotation = membrane_annotation, cell_data = cell_spatial,
    markers = unique(unlist(source_marker_sets)), ncol = 2))
}
```

### Membrane Cleanup Diagnostics

The correction step below applies all membrane-derived rules across all
annotated cell types. This creates a corrected run and then collects the
corrected cell-by-gene count matrix that will be used for downstream diagnostics.

```{r membrane-cleanup}
membrane_correction <- membrane_score$correct(rules = membrane_rules,
  name = "membrane_clean")
membrane_counts <- membrane_correction$counts()

membrane_correction$summary()
```

The next step is illustrative: repeat the same tumor-versus-adjacent-normal DE
comparison on the corrected counts, limited to the two diagnostic cell types
defined above. This lets us inspect how membrane-based cleanup changes the
example DE signal.

```{r membrane-diagnostic-de}
membrane_de <- celladmix_de_by_group(membrane_counts, cell_spatial,
  subset_by = "merged_annotation", subsets = diagnostic_cell_types,
  group_by = "domain_label", contrast = c("tumor", "adjacent_normal"))
```

This boxplot summarizes the number and fraction of molecules removed per cell.
It is a coarse sanity check: very large shifts indicate aggressive cleanup, while
near-zero shifts indicate that few rules applied to the dataset.

```{r membrane-removal, fig.width=6.5, fig.height=5.2}
membrane_correction$plot_removed_molecules()
```

The marker-expression plots compare source-marker expression in the original and
corrected count matrices. Here they are restricted to the two diagnostic target
cell types, so the plot illustrates cleanup impact rather than defining which
cell types were corrected.

```{r membrane-marker-expression, fig.width=11.0, fig.height=4.2}
plots <- lapply(diagnostic_cell_types, function(cell_type) {
  cells <- cell_spatial$cell_id[cell_spatial$merged_annotation == cell_type]
  celladmix_plot_marker_expression(original_counts, membrane_counts,
    cells = cells, markers = source_marker_sets) +
    ggplot2::ggtitle(cell_type)
})
cowplot::plot_grid(plotlist = plots, ncol = length(plots), align = "hv", axis = "tblr")
```

The volcano plots show the same tumor-versus-adjacent-normal DE comparison
before and after cleanup. The goal is to check whether highlighted source-marker
genes become less prominent after removing putative admixture molecules.

```{r membrane-volcano, fig.width=11.0, fig.height=4.8}
for (target_type in diagnostic_cell_types) {
  p_before <- celladmix_plot_volcano(diagnostic_de_original[[target_type]],
    markers = source_marker_sets, title = paste(target_type, "before cleanup")) +
    ggplot2::theme(legend.position = "none")
  p_after <- celladmix_plot_volcano(membrane_de[[target_type]],
    markers = source_marker_sets, title = paste(target_type, "after membrane cleanup"))
  print(cowplot::plot_grid(plotlist = list(p_before, p_after),
    ncol = 2, align = "hv", axis = "tblr"))
}
```

The scatter plots below compare each gene's adjusted DE significance before and
after cleanup. Genes below the diagonal are less significant after correction;
highlighted markers show whether expected source signatures are reduced in the
diagnostic target cell types.

```{r membrane-pvalue-shift, fig.width=11.0, fig.height=5.0}
plots <- lapply(diagnostic_cell_types, function(target_type) {
  celladmix_plot_de_shift(diagnostic_de_original[[target_type]], membrane_de[[target_type]],
    markers = source_marker_sets, title = paste("Membrane", target_type, "p-value shift"))
})
cowplot::plot_grid(plotlist = plots, ncol = length(plots), align = "hv", axis = "tblr")
```

## Bridge Scoring

Bridge scoring uses molecule adjacency across candidate cell pairs rather than
the stain image. It is useful when a source cell is visible and physically
adjacent to target molecules.

```{r bridge-score}
bridge_score <- fit$score_bridge()
bridge_annotation <- bridge_score$annotation(p_thresh = p_thresh)
bridge_rules <- bridge_score$rules(p_thresh = p_thresh)

head(bridge_rules[, c("factor", "source_cell_type", "target_cell_type", "p_value")])
```

```{r bridge-heatmap, fig.width=8.5, fig.height=5.4}
bridge_score$plot_heatmap(p_thresh = p_thresh,
  source_prior = factor_source_score)
```

```{r bridge-score-pairs, fig.width=score_pairs_width, fig.height=score_pairs_height}
bridge_score$plot_pairs()
```

The correction step below applies all bridge-derived rules across all annotated
cell types. As above, the downstream diagnostic plots focus only on the
illustrative endothelial and immune target cells.

```{r bridge-cleanup}
bridge_correction <- bridge_score$correct(rules = bridge_rules,
  name = "bridge_clean")
bridge_counts <- bridge_correction$counts()
bridge_de <- celladmix_de_by_group(bridge_counts, cell_spatial,
  subset_by = "merged_annotation", subsets = diagnostic_cell_types,
  group_by = "domain_label", contrast = c("tumor", "adjacent_normal"))

bridge_correction$summary()
```

```{r bridge-removal, fig.width=6.5, fig.height=5.2}
bridge_correction$plot_removed_molecules()
```

```{r bridge-marker-expression, fig.width=11.0, fig.height=4.2}
plots <- lapply(diagnostic_cell_types, function(cell_type) {
  cells <- cell_spatial$cell_id[cell_spatial$merged_annotation == cell_type]
  celladmix_plot_marker_expression(original_counts, bridge_counts,
    cells = cells, markers = source_marker_sets) +
    ggplot2::ggtitle(cell_type)
})
cowplot::plot_grid(plotlist = plots, ncol = length(plots), align = "hv", axis = "tblr")
```

```{r bridge-volcano, fig.width=11.0, fig.height=4.8}
for (target_type in diagnostic_cell_types) {
  p_before <- celladmix_plot_volcano(diagnostic_de_original[[target_type]],
    markers = source_marker_sets, title = paste(target_type, "before cleanup")) +
    ggplot2::theme(legend.position = "none")
  p_after <- celladmix_plot_volcano(bridge_de[[target_type]],
    markers = source_marker_sets, title = paste(target_type, "after bridge cleanup"))
  print(cowplot::plot_grid(plotlist = list(p_before, p_after),
    ncol = 2, align = "hv", axis = "tblr"))
}
```

```{r bridge-pvalue-shift, fig.width=11.0, fig.height=5.0}
plots <- lapply(diagnostic_cell_types, function(target_type) {
  celladmix_plot_de_shift(diagnostic_de_original[[target_type]], bridge_de[[target_type]],
    markers = source_marker_sets, title = paste("Bridge", target_type, "p-value shift"))
})
cowplot::plot_grid(plotlist = plots, ncol = length(plots), align = "hv", axis = "tblr")
```

## Score Comparison

These summaries compare the source-cell annotation and target-cell significance
calls across membrane and bridge scoring. In each triangle panel, the upper-right
half of a cell corresponds to the first method named in the subtitle and the
lower-left half corresponds to the second.

```{r scoring-correspondence-plot, fig.width=10.0, fig.height=5.0}
celladmix_plot_score_agreement(list(Membrane = membrane_annotation,
  Bridge = bridge_annotation), p_thresh = p_thresh)
```
