---
title: "Generative Admixture Correction on Xenium Pancreas"
output:
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---

```{r setup, include=FALSE}
library(cellAdmixCore)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE,
  fig.align = "center", dpi = 96, fig.retina = 2)
```

Molecules leak between neighboring cells, so a cell's counts mix its own
expression with material from the cells around it. But not every count
that rises near a neighbor is leakage: cells also change their own
transcription in response to their neighborhood, and a correction that
flattens every neighbor-linked gradient erases exactly that biology. The
generative model separates the two: it decomposes each cell's counts into
the cell's own expression programs, contamination from each neighboring
cell type, ambient background, and neighborhood-induced expression, then
removes the contamination and ambient shares while keeping the induced
ones ([docs/generative.md](../../docs/generative.md) describes the model
and its validation; [the minimal version](pancreas_generative_minimal.md)
of this page is four calls long).

This example characterizes both sides on the pancreas dataset: how much
material moves between which cell types, and which genes the model
recognizes as the cells' own induced expression.

```{r fit}
annotation <- read.csv(file.path("annotations", "annotation.csv.gz"))
cell_annotation <- setNames(annotation$merged_annotation, annotation$cell_id)

ds <- cellAdmix("data", output_dir = "out", annotation = cell_annotation)
fit <- ds$fit(nmf_variant = "invsqrt_kl")
```

## The admixture pattern between cell types

The audit measures, for every ordered pair of cell types, how many
molecules leaked from the source type into cells of the target type,
using the rise of source-marker content with the number of source-type
neighbors:

```{r audit-map, fig.width=6.5, fig.height=5}
audit <- fit$audit_admixture()
audit$plot_map()
```

The magnitudes are far from uniform. Sorting the detected pairs by their
estimated admixture rate — the fraction of the target type's molecules
that originated in the source:

```{r audit-pairs}
pairs <- audit$pairs(detected_only = TRUE)
head(pairs[order(-pairs$rate),
  c("source", "target", "admixed_molecules", "rate")], 8)
```

Aggregating over targets shows which cell types shed the most material —
on this tissue, the exocrine compartment dominates as a source:

```{r audit-sources, fig.width=6, fig.height=3.2}
by_source <- aggregate(admixed_molecules ~ source, pairs, sum)
ggplot(by_source, aes(reorder(source, admixed_molecules),
    admixed_molecules / 1e3)) +
  geom_col(fill = "#c0392b", alpha = 0.85) +
  coord_flip() +
  labs(x = NULL, y = "estimated admixed molecules shed (thousands)") +
  theme_classic(base_size = 11)
```

## Correcting with the generative model

One call fits the model on all detected pairs (about twenty seconds on
this dataset):

```{r correct}
correction <- audit$correct_generative(num_threads = 8)
correction
```

The model's central quantity is the per-cell contamination fraction: how
much of each cell's content arrived from each source. Its dependence on
the number of source neighbors, for the largest pair:

```{r composition, fig.width=6, fig.height=3.6}
comp <- correction$composition("Exocrine epithelial",
  "Ductal/tumor epithelial")
cells_xy <- fit$cell_factors()
expo <- setNames(
  cellAdmixCore:::.celladmix_source_exposure_counts(
    cells_xy, cell_annotation, 15L)$counts[, "Exocrine epithelial"],
  as.character(cells_xy$cell_id))
comp$exposure <- pmin(expo[comp$cell_id], 4)
ggplot(comp, aes(factor(exposure), contamination)) +
  geom_boxplot(outlier.size = 0.3, fill = "#f4a582") +
  labs(x = "exocrine cells among 15 nearest neighbors",
    y = "fraction of the ductal cell's molecules\nattributed to exocrine contamination") +
  theme_classic(base_size = 11)
```

The same fractions drawn in space show the admixture concentrating at
the tissue interfaces:

```{r composition-map, fig.width=7, fig.height=5}
xy <- cells_xy[match(comp$cell_id, as.character(cells_xy$cell_id)), ]
ggplot(cbind(comp, x = xy$x, y = xy$y), aes(x, y,
    color = pmin(contamination, 0.6))) +
  geom_point(size = 0.25) +
  scale_color_viridis_c(option = "inferno",
    name = "exocrine share\nof ductal-cell\ncontent") +
  coord_equal() +
  labs(x = NULL, y = NULL,
    title = "Exocrine contamination of ductal cells") +
  theme_void(base_size = 11)
```

## The induction pattern

The model retains genes whose exposure-linked excess exceeds what
transfer from the source can deliver — expression the target cells
switched on themselves. On this tissue the retained set is led by the
duct-cell program that ductal cells activate at acinar interfaces:

```{r induced}
induced <- correction$induced
head(induced[order(-induced$excess),
  c("source", "target", "gene", "excess", "fold", "z")], 8)
```

The distinction is visible gene by gene. AMY2A is the largest exocrine
transfer channel: its rise with exocrine exposure in ductal cells is
transferred material, and the correction flattens it to the unexposed
level. CFTR carries an induced duct-cell program: its rise is the ductal
cells' own expression, and the correction preserves it while removing
only the transferred share:

```{r gene-contrast, fig.width=8.5, fig.height=3.6}
before <- fit$counts()
after <- correction$counts()
t_cells <- comp$cell_id
tot <- Matrix::colSums(before[, t_cells])
bins <- pmin(expo[t_cells], 3)
rate_by_bin <- function(m, gene) {
  tapply(m[gene, t_cells], bins, sum) / tapply(tot, bins, sum) * 1e3
}
df <- do.call(rbind, lapply(c("AMY2A", "CFTR"), function(g) {
  rbind(
    data.frame(gene = g, counts = "observed", bin = 0:3,
      rate = as.numeric(rate_by_bin(before, g))),
    data.frame(gene = g, counts = "corrected", bin = 0:3,
      rate = as.numeric(rate_by_bin(after, g))))
}))
ggplot(df, aes(bin, rate, color = counts)) +
  geom_line() + geom_point() +
  facet_wrap(~gene, scales = "free_y") +
  scale_color_manual(values = c(observed = "#c0392b",
    corrected = "#2980b9")) +
  labs(x = "exocrine cells among 15 nearest neighbors",
    y = "rate in ductal cells (per 1,000 molecules)") +
  theme_classic(base_size = 11)
```

## Verification

The correction is verified against the audit's own measurements, like
any other correction in the package:

```{r verify, fig.width=8, fig.height=4}
report <- audit$evaluate(correction)
report$summary()
report$plot_cleanup()
```

Two limits are worth keeping in mind when reading the results. The model
never removes a target type's own marker genes (their contamination is
indistinguishable from native expression and is deliberately left in
place), so shared-gene admixture is under-removed by construction. And
induced expression is only recognizable when it is disproportionate to
the source's transfer profile: induction of a source's own top marker
genes cannot be told apart from admixture at count level and is removed
with it ([docs/generative.md](../../docs/generative.md), the
identifiability limit).
