---
title: "cellAdmix with a Seurat Xenium Object"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE,
  fig.width = 8, fig.height = 6, dpi = 192)

library(Seurat)
library(cellAdmixCore)
library(ggplot2)
library(cowplot)
```

## Overview

This example shows how to use cellAdmix when the cell-level analysis is managed
in Seurat. Seurat defines the active cells, active genes, cell annotations, and
downstream object that users keep working with. cellAdmix uses the original
Xenium bundle as a molecule-complete backing source, fits molecular admixture
factors, scores likely source/target relationships, removes putative admixture
molecules, and writes corrected counts back into the Seurat object as a new
assay.

The core workflow is:

```{r api-overview, eval=FALSE}
seu <- LoadXenium("data", molecule.coordinates = FALSE)
seu$cell_type <- ...

ds <- cellAdmix(seu, xenium_dir = "data", output_dir = "out",
  annotation = "cell_type")
fit <- ds$fit()
score <- fit$score_membrane()
correction <- score$correct()

seu <- celladmix_add_to_seurat(seu, fit = fit, score = score,
  correction = correction, corrected_assay = "cellAdmix")
```

We use `molecule.coordinates = FALSE` when loading Seurat because cellAdmix reads
the complete transcript table directly from the Xenium bundle. This avoids
duplicating all molecules in R memory and preserves Xenium fields that Seurat's
molecule object does not keep.

## Input Data

This example uses the 10x Genomics
[FFPE human pancreas Xenium dataset](https://www.10xgenomics.com/datasets/ffpe-human-pancreas-with-xenium-multimodal-cell-segmentation-1-standard).
Download and unpack the bundle into a folder named `data` next to this notebook:

```{bash download-data, eval=FALSE}
curl -O https://cf.10xgenomics.com/samples/xenium/2.0.0/Xenium_V1_human_Pancreas_FFPE/Xenium_V1_human_Pancreas_FFPE_outs.zip
unzip Xenium_V1_human_Pancreas_FFPE_outs.zip
mv outs data
```

The quick cell-type annotations used here are for demonstration only. They were
derived from a lightweight clustering/marker workflow in the companion pancreas
annotation notebook and are intentionally not treated as a curated reference.
In a normal Seurat analysis, these labels would usually come from standard
Seurat clustering, marker inspection, reference mapping, or another annotation
workflow. Here we load predetermined labels only to keep the example focused on
cellAdmix integration. Place `annotation.csv.gz` in an `annotations` folder next
to this notebook.

```{r paths}
data_dir <- "data"
annotation_file <- file.path("annotations", "annotation.csv.gz")

stopifnot(dir.exists(data_dir))
stopifnot(file.exists(annotation_file))
```

## Load Xenium into Seurat

We first create a normal Seurat object from the Xenium output. This is where a
typical Seurat workflow would do QC, subsetting, normalization, clustering, and
manual or automated annotation. In this tutorial, we keep that part minimal and
attach the precomputed example annotations.

The annotation file is a gzip-compressed CSV with one row per cell. We use
`cell_id` to match rows to Seurat cells, and `merged_annotation` as the
cell-type label that will be stored in `seu$cell_type`.

```{r load-seurat}
seu <- LoadXenium(data_dir, fov = "fov", assay = "Xenium",
  molecule.coordinates = FALSE)

annotation <- read.csv(gzfile(annotation_file), stringsAsFactors = FALSE)
rownames(annotation) <- annotation$cell_id
seu$cell_type <- annotation[colnames(seu), "merged_annotation"]
seu <- subset(seu, cells = colnames(seu)[!is.na(seu$cell_type)])

table(seu$cell_type)
```

The Seurat object now defines the active analysis universe. If the object had
already been subsetted to particular cells or genes, the cellAdmix adapter would
pass those active cell and feature IDs as filters to the Xenium-backed store.

```{r seurat-spatial-labels, fig.width=7, fig.height=6}
ImageDimPlot(seu, fov = "fov", group.by = "cell_type",
  axes = TRUE, cols = "polychrome", dark.background = FALSE,
  border.color = "grey30", border.size = 0.03, alpha = 0.9) +
  ggtitle("Cell annotations")
```

## Construct cellAdmix Dataset

The constructor below is the main integration point. The first argument is the
Seurat object, while `xenium_dir` points to the original Xenium bundle. The
annotation argument names a Seurat metadata field. We do not need to specify
`assay` or `image` here because the defaults are inferred from the Seurat object.
The output directory stores the cellAdmix project cache: encoded molecules,
factorization output, scores, correction results, and any reusable intermediate
files.

cellAdmix can work with Seurat without loading molecule coordinates into R
memory, but it still needs a molecule-complete backing source for the core
fit/score/correct workflow. For Xenium data, that source is the original bundle
passed as `xenium_dir`; without it, the analysis is limited to Seurat-side
metadata/visualization or adding previously computed cellAdmix results back to
the object.

```{r construct-dataset}
output_dir <- "out_seurat"
num_threads <- min(10, parallel::detectCores())

ds <- cellAdmix(seu, xenium_dir = data_dir, output_dir = output_dir,
  annotation = "cell_type", num_threads = num_threads)

ds
```

## Fit Molecular Factors

`fit()` samples neighborhood composition vectors from molecules inside cells and
fits NMF factors. The default rank is chosen from the number of annotation
classes. Multiple random/cluster-initialized NMF runs are run in parallel by
default, and the lowest-loss solution is retained.

```{r fit}
fit <- ds$fit(verbose = TRUE)
fit
```

Top factor-loading genes help interpret each factor. The stability plot shows
whether important factors were consistently recovered across NMF restarts.
Factor importance (the x-axis, echoed by dot size) is the share of all assigned molecules that carry the factor's label, so large dots mark factors that dominate the tissue-wide signal.

```{r factor-loadings, fig.width=10, fig.height=7}
fit$plot_loadings()
```

```{r factor-stability, fig.width=6, fig.height=5}
fit$plot_stability()
```

## Membrane-Based Scoring

This Xenium run includes membrane staining, so we use membrane scoring. The
score summarizes which factor-cell-type combinations look more consistent with
external admixture than with native signal.

```{r membrane-score}
membrane_score <- fit$score_membrane(verbose = TRUE)
```

The heatmap is a compact factor-by-cell-type summary of putative source and
admixture labels.

```{r membrane-heatmap, fig.width=8, fig.height=6}
p_thresh <- 0.1
membrane_score$plot_heatmap(p_thresh = p_thresh)
```

The pair plots show, for each factor, which target cell types have significant
evidence of admixture from each candidate source cell type. The source evidence
bar above each matrix summarizes which cell types best explain the factor.

```{r membrane-pairs, fig.width=9.6, fig.height=20}
membrane_score$plot_pairs()
```

Correction rules are the source-factor-target combinations that pass the
threshold and will be used to remove molecules from target cells.

```{r membrane-rules}
membrane_rules <- membrane_score$rules(p_thresh = p_thresh)
head(membrane_rules[, c("factor", "source_cell_type", "target_cell_type", "p_value")])
```

## Correct Counts

The correction step removes molecules assigned to significant admixture factors
in their target cell types. The result is persisted on disk and can be collected
as a sparse gene-by-cell matrix.

```{r correction}
membrane_correction <- membrane_score$correct(rules = membrane_rules,
  name = "membrane_clean")

membrane_correction$summary()
```

The molecule-removal plot is a basic diagnostic for how much correction was
applied across cell types. Large removals should be inspected carefully.

```{r molecule-removal, fig.width=8, fig.height=6}
membrane_correction$plot_removed_molecules()
```

## Add Results Back to Seurat

Finally, we attach cellAdmix outputs to the Seurat object. Factor fractions and
correction summaries are added to metadata. Corrected counts are added as a new
`cellAdmix` assay, leaving the original `Xenium` assay unchanged.

```{r add-to-seurat}
seu <- celladmix_add_to_seurat(seu, fit = fit, score = membrane_score,
  correction = membrane_correction, corrected_assay = "cellAdmix")

SeuratObject::Assays(seu)
grep("^celladmix_", colnames(seu[[]]), value = TRUE)
dim(seu[["cellAdmix"]])
```

At this point, downstream Seurat analyses can compare the original `Xenium`
assay with the corrected `cellAdmix` assay, or use the added cellAdmix metadata
for factor-aware visualization and diagnostics.
