---
title: "cellAdmix Quickstart: Admixture Correction on Xenium Pancreas"
output:
  html_document:
    toc: false
---

```{r setup, include=FALSE}
library(cellAdmixCore)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE,
  fig.align = "center", dpi = 192)
```

This is the compact version of the recommended cellAdmix workflow:
estimate how much admixture the dataset carries, correct it, and verify
the result. Inputs are the Xenium output bundle (`data/`) and a cell-type
annotation (see the
[full tutorial](pancreas_membrane_scoring_clean.ipynb) for download
instructions and a walkthrough of the intermediate diagnostics; see
[docs/benchmarks.md](../../docs/benchmarks.md) for how the settings below
were chosen).

```{r construct}
annotation <- read.csv(file.path("annotations", "annotation.csv.gz"))
cell_annotation <- setNames(annotation$merged_annotation, annotation$cell_id)

ds <- cellAdmix("data", output_dir = "out", annotation = cell_annotation)
fit <- ds$fit(nmf_variant = "invsqrt_kl")
```

For membrane-stained data, membrane scoring with the `invsqrt_kl`
factorization removes the most admixture in our benchmarks; on data
without stains, use `fit$score_bridge()` with the default `ls_nmf`
factorization instead.

## How much admixture is there?

The audit estimates, for every ordered cell-type pair, how many molecules
leaked from the source type into target cells — using only the spatial
exposure structure, before any correction:

```{r audit-map, fig.width=6.8, fig.height=5.2}
audit <- fit$audit_admixture()
audit$plot_map()
```

## Correct

One quick check that the factorization is sound — most factors should sit
above the 0.3 stability threshold:

```{r stability, fig.width=6.2, fig.height=4.2}
fit$plot_stability()
```

Scoring turns factors into removal rules (each vetted by a native-factor
false-positive check), and correction removes the matching molecules:

```{r correct}
score <- fit$score_membrane()
correction <- score$correct(rules = score$rules())
```

## Verify

The audit re-measures the same exposure gradients on the corrected counts.
The report gives per-pair cleanup sensitivity, the own-marker
false-removal rate (how much almost-surely-genuine expression was lost),
and warns about any detected pair the rules did not cover:

```{r verify, warning=TRUE, fig.width=9.6, fig.height=4.0}
report <- audit$evaluate(correction)
report$summary()

cowplot::plot_grid(
  audit$plot_remaining(list(corrected = correction)),
  audit$plot_exposure(correction = correction),
  ncol = 2, align = "hv", axis = "tblr")
```

## Use the corrected counts

```{r export}
corrected_counts <- correction$counts()
dim(corrected_counts)
```

`corrected_counts` is a sparse genes × cells matrix ready for downstream
analysis; `celladmix_add_to_seurat()` and the SpatialData helpers attach it
to existing objects.
