---
title: "Generative Admixture Correction in Brief: Xenium Pancreas"
output:
  github_document:
    toc: false
    html_preview: false
---

```{r setup, include=FALSE}
library(cellAdmixCore)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE,
  fig.align = "center", dpi = 96, fig.retina = 2)
```

The generative model corrects admixture by decomposing each cell's counts
into the cell's own expression, contamination from each neighboring cell
type, ambient background, and expression induced by the neighborhood —
removing the contamination and ambient shares while keeping the induced
ones. This page shows the shortest path from data to corrected counts;
the [full example](pancreas_generative.md) characterizes what the model
measures along the way, and [docs/generative.md](../../docs/generative.md)
describes the model itself.

```{r construct}
annotation <- read.csv(file.path("annotations", "annotation.csv.gz"))
cell_annotation <- setNames(annotation$merged_annotation, annotation$cell_id)

ds <- cellAdmix("data", output_dir = "out", annotation = cell_annotation)
```

The audit measures the admixture straight from the dataset — no
factorization is involved. The generative model is then fitted on its
detected pairs; the NMF fit enters only to initialize the model's
expression programs, the recommended configuration (`init = "clusters"`
fits without it):

```{r correct}
audit <- ds$audit_admixture()
nmf_fit <- ds$fit(nmf_variant = "invsqrt_kl")
model <- audit$fit_generative(init = nmf_fit, num_threads = 8)
model
correction <- model$correct()
```

The corrected gene-by-cell matrix drops into any downstream analysis:

```{r counts}
corrected_counts <- correction$counts()
dim(corrected_counts)
```

The model keeps the genes whose exposure-linked excess reflects the
cells' own response to their neighbors rather than transferred material:

```{r induced}
head(model$induced[order(-model$induced$excess), ], 5)
```

And it is verified like any other correction, against the audit's own
measurements:

```{r verify}
report <- audit$evaluate(correction)
report$summary()
```
